---
title: "Lab 01a"
author: Dr. Medina
format: html
knitr:
  opts_chunk:
    message: false
    warning: false
---

## Preface 

This file is a **Quarto document** and it allows us to combine text and code.
By clicking **Render**, an html will be produced with all of the output from our code included.
For today we will not worry about how the code, instead we will focus on our investigation.

Start by clicking `Render` and examine the html produced.

In this lab we will explore popular baby names for the state of California.
This data comes from California birth certificates and is available on the [California Open Data Portal](https://data.ca.gov/dataset/most-popular-baby-names).

## Obtain and familiarize yourself with the data

The functions we will use to explore the data are contained in **packages**.
In order to use these functions we have to load the packages into our session environment.
```{r}
library(tidyverse)
```

This data is read in from a file called `california-popular-baby-names.csv` available online.
We will save this data into an **object** called `baby-names`.
```{r}
#baby_names <- read_csv("data/california-popular-baby-names.csv")

baby_names <- read_csv("csuci-data200.github.io/data200-week01/data/california-popular-baby-names.csv")
```

Let's look at what is in our `baby_names` object.
```{r}
baby_names
```

This dataset has 5 columns and 3,203 rows in total, which would be a lot to print, so only the first 10 are displayed.


For each year and specified sex at birth this data contains the 25 most popular baby first names and counts.
We have data for each year from 1960 to 2023.

Let's use this data to explore some questions.

## What were the 20 most popular baby names in 2023?

```{r}
baby_names |> 
  filter(Year == 2023) |>
  arrange(desc(Count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = Count, y = reorder(Name, Count))) +
  geom_count() +
  geom_text(aes(label = Count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born in 2023",
    y = "Name"
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))
```

## What were the 20 most popular baby names in the year you were born?

In the Quarto document change the "2023" below to the year you were born and render the html again.
```{r}
baby_names |> 
  filter(Year == 2023) |>
  arrange(desc(Count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = Count, y = reorder(Name, Count))) +
  geom_count() +
  geom_text(aes(label = Count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born in 2023",
    y = "Name"
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))
```
## What are the top 20 most popular baby names overall from all babies born between 1960-2023?

```{r}
all_names_plot <- baby_names |>
  group_by(Name) |> 
  summarise(total_count = sum(Count)) |> 
  arrange(desc(total_count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = total_count, y = reorder(Name, total_count))) +
  geom_count() +
  geom_text(aes(label = total_count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born from 1960-2023",
    y = "Name"
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))

all_names_plot
```

## Is your name in the list of the most popular baby names?

```{r}
sort(unique(baby_names$Name))
```



## Are the top 20 most popular baby names overall the same when we correct for truncation?

If you look into the **data dictionary** it says:
*Names in records from data years 1960 through 1977 are truncated to 8 characters, and 1978 through 2017 are truncated to 12 characters.* 
*For example, the names CHRISTINA, CHRISTINE, and CHRISTIN would all be included in CHRISTIN*.*

When we look at names across the entire time period we may get different results if we truncated all names to 8 characters.
This truncation is accounted for in the plot below.

```{r}
baby_names <- baby_names |> 
  mutate(name_fixed = case_when(
    str_length(Name) > 8 ~ (str_trunc(Name, width = 9, ellipsis = "*")),
    str_length(Name) <= 8 ~ Name
  )) 

corrected_all_names_plot <-  baby_names |> 
  group_by(name_fixed) |> 
  summarise(total_count = sum(Count)) |> 
  arrange(desc(total_count)) |> 
  slice(1:20) |> 
  ggplot(aes(x = total_count, y = reorder(name_fixed, total_count))) +
  geom_count() +
  geom_text(aes(label = total_count), hjust = -0.25) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    x = "Number of babies born from 1960-2023",
    y = "Corrected names"
  )  +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.15)))

gridExtra::grid.arrange(all_names_plot, corrected_all_names_plot, ncol = 2)
```

Do you notice any difference?


## How did the 5 overall most popular names change over the years?

```{r}
most_popular_names <- baby_names |> 
  group_by(name_fixed) |> 
  summarise(total_count = sum(Count)) |> 
  arrange(desc(total_count)) |> 
  slice(1:5) |> 
  pull(name_fixed)

baby_names |> 
  filter(name_fixed %in% most_popular_names) |> 
  ggplot(aes(x = Year, y = Count, color = name_fixed, linetype = name_fixed)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Number of babies born",
    color = "Name",
    linetype = "Name"
  ) +
  ylim(0, 9000)
```

Did the most popular name in 1960 stay the most popular? 

If not, when did a different name become more popular and what was it?

Notice that the line for the name "Matthew" does not start until approximately the year 1968. Why do you think that is?
