---
title: "Tidyverse Style Guide"
format: html
execute: 
  eval: false
---

The full Tidyverse Style Guide is available at <https://style.tidyverse.org/>.

We have not covered all the content in the style guide, but through different lectures we have learned the style rules we are following.

The major rules fall under three tasks, naming objects, spacing/indenting code, and naming files.

## Naming objects

- Variable and object names should use only lowercase letters, numbers, and _. Use underscores (`_`) (so called snake case) to separate words within a name.

```{r}
# Good
day_one
day_1

# Bad
DayOne
dayone
```

## Spacing and indenting

- Most operators (==, +, -, <-, *, etc.) should always be surrounded by spaces.

```{r}
# Good
height <- (feet * 12) + inches
mean(x, na.rm = TRUE)

# Bad
height<-feet*12+inches
mean(x, na.rm=TRUE)
```

- If a function call is too long to fit on a single line, use one line each for the function name, each argument, and the closing ). This makes the code easier to read and to change later.

```{r}
# Good
mutate(
  penguins, 
  species_abbreviated = case_when(
    species == "Gentoo" ~ "G",
    species == "Chinstrap" ~ "C",
    specied == "Adelie" ~ "A"
  )
)

# Bad
mutate(penguins, species_abbreviated = case_when(species == "Gentoo" ~ "G", species == "Chinstrap" ~ "C",
  specied == "Adelie" ~ "A"
))
```

- A pipe `|>` should be followed by a function on a new line, not the same line.

```{r}
# Good
new_penguins_data <- penguins |> 
  filter(islans != "Dream") |> 
  mutate(species_abbreviated = case_when(
    species == "Gentoo" ~ "G",
    species == "Chinstrap" ~ "C",
    specied == "Adelie" ~ "A"
  ))

# Bad
new_penguins_data <- penguins |> filter(islans != "Dream") |> mutate(species_abbreviated = case_when(species == "Gentoo" ~ "G",
  species == "Chinstrap" ~ "C",
  specied == "Adelie" ~ "A"
))
```

- The `+` operator for ggplots should be followed by a function on a new line, not the same line.

```{r}
# Good
ggplot(penguins, aes(x = bill_len, y = bill_dep)) +
  geom_point()

# Bad
ggplot(penguins, aes(x = bill_len, y = bill_dep)) + geom_point()
```

## Naming files

- File names should use only lowercase letters, numbers, and -. Use dashes (-) to separate words within a name. For example`tidyverse-style-guide-summary.qmd` or `popular-baby-names.csv`. This is only relevant when you are creating a file or exporting data to a file.
