---
title: "Data 200: Lecture 8b"
subtitle: "Frequentist inference"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

## 

```{r}
#| echo: false
options(scipen=0)
library(bayesrules)
library(tidyverse)
library(scales)
```

```{r}
#| echo: false
theme_set(theme_bw(base_size = 18))
```

# Central Limit Theorem

##

|                               | Parameter of Interest | Point Estimate / Sample Statistic |
|-------------------------------|-----------------------|-----------------------------------|
| Mean                          | $\mu$                 | $\bar x$                          |
| Difference of Two Means       | $\mu_1 - \mu_2$       | $\bar x_1 - \bar x_2$             |
| Proportion                    | $\pi$                 | $p$                               |
| Difference of Two Proportions | $\pi_1 - \pi_2$       | $p_1 - p_2$                       |

## 
```{r fig.height = 4, fig.align='center'}
#| echo: false
set.seed(7)

bike_prop <- 0.15

bike <- c(rep("yes", bike_prop*10000), rep("no", (1-bike_prop)*10000))

pop <- tibble(bike)


pop %>% 
  ggplot() +
  aes(x = bike, y = ..prop.., group = 1) +
  geom_bar() +
  labs(title = "Population Distribution")
```

Let $\pi$ represent the proportion of bike owners on campus then $\pi =$ `r bike_prop`. 

Let's pretend we have never seen the data and we do not know the population parameter $\pi$. In fact this is usually what happens in real life. 

- We do not have the population information but we do want to know a population __parameter__ (does not necessarily have to be the proportion).

. . .

- If we took a sample and calculated the sample mean, we would name this __point estimate__ of the parameter. 


## Getting to sampling distribution of a proportion

$p_1$ - Proportion of first sample (n = 100)

```{r}
sample1 <- pop %>% 
  sample_n(100) %>% 
  count(bike) %>% 
  mutate(prop = n/sum(n)) %>% 
  filter(bike == "yes") %>% 
  select(prop) %>% 
  pull()

sample1
```

## Getting to sampling distribution of a proportion

$p_2$ -Proportion of second sample (n = 100)

```{r}
pop %>% 
  sample_n(100) %>% 
  count(bike) %>% 
  mutate(prop = n/sum(n)) %>% 
  filter(bike == "yes") %>% 
  select(prop) %>% 
  pull()
```

## Getting to sampling distribution of a proportion

$p_3$ -Proportion of third sample (n = 100)

```{r}
pop %>% 
  sample_n(100) %>% 
  count(bike) %>% 
  mutate(prop = n/sum(n)) %>% 
  filter(bike == "yes") %>% 
  select(prop) %>% 
  pull()
```

## Sampling Distribution of a Proportion across 10,000 samples

```{r cache = TRUE}
#| echo: false
sample_props <- vector()


for (i in 1:10000){
 
    sample <- pop %>% 
    sample_n(100) 
    
     sample_props[i] <- min(prop.table(table(sample$bike)))
  
  
  
}

sample_props_data <- tibble(sample_props = sample_props)
```

```{r}
#| echo: false
sample_props_data %>% 
  ggplot() +
  aes(x = sample_props) +
  geom_histogram(binwidth=0.01) +
  geom_vline(xintercept = bike_prop, color= "#e56646") +
  labs(x = "Sample proportions")
```

##

If certain conditions are met then

$$p \sim \text{approximately } N(\pi, \frac{\pi(1-\pi)}{n})$$

##

When certain conditions are met then:

$$p \sim  \text{approximately } N(\pi,  \frac{\pi(1-\pi)}{n})$$

. . .

$$(p_1 - p_2) \sim \text{approximately } N((\pi_1 - \pi_2),  {\frac{\pi_1(1-\pi_1)}{n_1} + \frac{\pi_2(1-\pi_2)}{n_2}})$$

. . .


$$\bar x \sim \text{approximately }N( \mu, \frac{\sigma^2}{n})$$

. . .

$$(\bar x_1 - \bar x_2) \sim \text{approximately } N(\mu_1 - \mu_2, \frac{\sigma_1^2}{n_1}+ \frac{\sigma_2^2}{n_2})$$



## Central Limit Theorem (CLT)

If certain conditions are met, the sampling distribution will be normally distributed with a mean equal to the population parameter. The standard deviation will be inversely proportional to the square root of the sample size. 

. . .

Moving forward we will use CLT to make __inference__ about population parameters using sample statistics. 

## Take-Away Messages

- Sample statistics are point estimates. They are not the same thing as population parameters. 

. . .

- Point estimates can vary from sample to sample. Sampling distribution captures this variance. 

. . .

- Sampling distribution is never observed. 



# Confidence Interval for a Proportion

```{r }
#| echo: false
set.seed(7)

bike_prop <- 0.15

bike <- c(rep("yes", bike_prop * 10000), rep("no", (1 - bike_prop) * 10000))

pop <- tibble(bike)
```

```{r cache = TRUE}
#| echo: false
sample_props <- vector()


for (i in 1:10000){
 
    sample <- pop %>% 
    sample_n(100) 
    
     sample_props[i] <- min(prop.table(table(sample$bike)))
  
  
  
}

sample_props_data <- tibble(sample_props = sample_props)
```

## In Reality

- We only have one sample and thus one point estimate of the population parameter. How can make use of it? 

. . .

- First we will assume the sample proportion is the best thing we have at hand and use it as a point estimate of the population proportion. 

. . .

- Second, even though we embrace the sample proportion as a point estimate of the population proportion, we will need to acknowledge that it has some error. 


## Standard Error


$p \sim  \text{approximately } N(\text{mean} = \pi, \text{sd} = \sqrt{\frac{\pi(1-\pi)}{n}})$

. . .

We call the standard deviation of the sampling distribution __standard error__ of the estimate. 

Standard error of a proportion is $\sqrt{\frac{p(1-p)}{n}}$.



## Confidence Interval 

CI = $\text{point estimate} \pm \text { margin of error}$

. . .

CI = $\text{point estimate} \pm \text { critical value} \times \text{standard error}$

. . .

CI for a proportion = $p \pm \text {critical value} \times \text{standard error}$

. . .

CI for a proportion = $p \pm \text {critical value} \times \sqrt{\frac{p(1-p)}{n}}$

. . .

95% CI for a proportion = $p \pm 1.96 \times \sqrt{\frac{p(1-p)}{n}}$ because ...

##

95% of the data falls within 1.96 standard deviations in the normal distribution.
      
```{r}
#| echo: false
z <- seq(-4, 4, by = 0.01)
y <- dnorm(z)
data <- tibble(z = z, y = y)
qplot(z, y, data = data, geom = "line") +
  geom_ribbon(
    data = subset(data, z > -1.96 & z <1.96),
    aes(ymax = y),
    ymin = 0,
    fill = "#e56646", 
    colour = NA, 
    alpha = 0.5
  ) +
  ylab("") +
  scale_x_continuous(breaks = c(-1.96, 0, 1.96)) +
  annotate(geom = "text", x = 0, y = 0.15, label = "95%") +
  annotate(geom = "text", x = 2.3, y = 0.01, label = "2.5%") +
  annotate(geom = "text", x = - 2.3, y = 0.01, label = "2.5%")

  
```
      

## How do we know that?

```{r echo = TRUE}
qnorm(0.025, mean = 0 , sd = 1)
```

```{r echo = TRUE}
qnorm(0.975, mean = 0 , sd = 1)
```


## 95% CI for the first sample

Recall $p = 0.17$ and $n = 100$

. . .

95% CI for a proportion = $p \pm 1.96 \times \sqrt{\frac{p(1-p)}{n}}$ 

. . .

95% CI = $0.17 \pm 1.96 \times \sqrt{\frac{0.17(1-0.17)}{100}}$ 

. . .

95% CI = $0.17 \pm 1.96 \times 0.03756328$ 

. . .

95%CI = $0.17 \pm 0.07362403$

. . .

95%CI = (0.09637597, 0.243624)


## 95% CI for the first sample

95%CI = (0.09637597, 0.243624)

We are 95% confident that the true population proportion of bike owners is in this confidence interval.


95%CI = (0.09637597, 0.243624)


```{r}
#| echo: false
tibble(p = bike_prop, n = 1, lower_bound = 0.09637597, upper_bound = 0.243624) %>% 
  ggplot() +
  aes(x = 0.17, y = n) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower_bound, xmax = upper_bound, height = .01)) +
  labs(y = " ", x = " ") +
  theme(axis.text.y = element_blank()) 
```


## Understanding Confidence Intervals


I have taken 100 samples with $n = 100$, calculated the sample proportion, standard error, and 95% CI interval for each sample

```{r}
#| echo: false
set.seed(26)

cv <- qnorm(0.025)

ci_data <- sample_props_data %>% 
  sample_n(100) %>% 
  rename(p = sample_props) %>% 
  mutate(
    SE = sqrt((p * (1 - p)) / 100),
    lower_bound = p + (cv * SE),
    upper_bound = p  - (cv * SE)
  )


ci_data
```


## Understanding Confidence Intervals

```{r}
#| echo: false
ci_data <-
  ci_data %>% 
  mutate(n = 1:nrow(ci_data)) %>% 
  mutate(contains = ifelse(lower_bound < bike_prop & upper_bound > bike_prop , 1,0)) 
```


```{r}
#| echo: false
ci_data %>% 
  ggplot() +
  aes(x = p, y = n) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower_bound, xmax = upper_bound)) +
  labs(y = " ") +
  theme(axis.text.y = element_blank()) 
```


## Understanding Confidence Intervals


```{r}
#| echo: false
ci_data %>% 
  ggplot() +
  aes(x = p, y = n, color = as.factor(contains)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower_bound, xmax = upper_bound)) +
  labs(y = " ") +
  theme(axis.text.y = element_blank(),
        legend.position = "none")  +
  geom_vline(xintercept =  bike_prop, color= "#e56646") 
  
```


## Confidence Interval Width

Which of the following confidence intervals would be the widest? Why?

- 90% CI
- 95% CI
- 99% CI

##

CI = $\text{point estimate} \pm \text { critical value} \times \text{standard error}$


```{r echo = TRUE}
qnorm(0.05, mean = 0, sd = 1) # critical value for 90% CI
```


```{r fig.height=4}
#| echo: false
z <- seq(-4, 4, by = 0.01)
y <- dnorm(z)
data <- tibble(z = z, y = y)
qplot(z, y, data = data, 
          geom = "line") +
  geom_ribbon(data = subset(data, z > -1.64 & z <1.64),
                  aes(ymax = y), 
                  ymin = 0,
                  fill = "#e56646", 
                  colour = NA, 
                  alpha = 0.5) +
  ylab("") +
  scale_x_continuous(breaks = c(-1.64,0,1.64)) +
  annotate(geom = "text", x = 0, y = 0.15, label = "90%") +
  annotate(geom = "text", x = 2, y = 0.03, label = "5%") +
  annotate(geom = "text", x = - 2, y = 0.03, label = "5%")

  
```

##

CI = $\text{point estimate} \pm \text { critical value} \times \text{standard error}$


```{r echo = TRUE}
qnorm(0.025, mean = 0, sd = 1) # critical value for 95% CI
```


```{r fig.height=4}
#| echo: false
z <- seq(-4, 4, by = 0.01)
y <- dnorm(z)
data <- tibble(z = z, y = y)
qplot(z, y, data = data, 
          geom = "line") +
  geom_ribbon(data = subset(data, z > -1.96 & z <1.96),
                  aes(ymax = y), 
                  ymin = 0,
                  fill = "#e56646", 
                  colour = NA, 
                  alpha = 0.5) +
  ylab("") +
  scale_x_continuous(breaks = c(-1.96,0,1.96)) +
  annotate(geom = "text", x = 0, y = 0.15, label = "95%") +
  annotate(geom = "text", x = 2.23, y = 0.01, label = "2.5%") +
  annotate(geom = "text", x = - 2.23, y = 0.01, label = "2.5%")

  
```

##

CI = $\text{point estimate} \pm \text { critical value} \times \text{standard error}$


```{r echo = TRUE}
qnorm(0.005, mean = 0, sd = 1) # critical value for 99%CI
```


```{r fig.height=4}
#| echo: false
z <- seq(-4, 4, by = 0.01)
y <- dnorm(z)
data <- tibble(z = z, y = y)
qplot(z, y, data = data, 
          geom = "line") +
  geom_ribbon(data = subset(data, z > -2.58 & z <2.58),
                  aes(ymax = y), 
                  ymin = 0,
                  fill = "#e56646", 
                  colour = NA, 
                  alpha = 0.5) +
  ylab("") +
  scale_x_continuous(breaks = c(-2.58,0,2.58)) +
  annotate(geom = "text", x = 0, y = 0.15, label = "99%")

  
```

##

CI = $\text{point estimate} \pm \text { critical value} \times \text{standard error}$

- 99% CI has the highest critical value.
- Higher critical value means higher margin of error.
- Higher margin of error means wider CI.

Thus 99% CI would be the widest.


## Effect of Sample Size on Confidence Interval 

Researchers A, B, and C are interested in proportion of bike ownership took samples. They each take separate samples of size 100, 500, and 1000 respectively. They each have a sample proportion of 0.18. What a surprise! Which of the researchers will find the narrowest 95% CI?

##

**Researcher A:** 95% CI for a proportion = $0.18 \pm 1.96 \times \sqrt{\frac{0.18(1-0.18)}{100}}$ 

**Researcher B:** 95% CI for a proportion = $0.18 \pm 1.96 \times \sqrt{\frac{0.18(1-0.18)}{500}}$ 

**Researcher C:** 95% CI for a proportion = $0.18 \pm 1.96 \times \sqrt{\frac{0.18(1-0.18)}{1000}}$ 

. . .

As sample size increases, the standard error decreases and the margin of error also decreases, thus the confidence interval interval gets narrower. The Researcher C would have the narrowest CI.


## CLT

If these conditions are met then

$p \sim \text{approximately } N(\pi, \frac{\pi(1-\pi)}{n})$


## Conditions


1. The sample data are independent. 
2. There needs to be at least 10 successes and 10 failures in the sample.
3. The sample size is smaller than 10% of the population.

##

__Example__

According to a Gallup Survey  of 1017 adults living in US 66% of Americans favor legalizing marijuana. 

Compute 95% confidence interval for the population proportion of those who favor legalizing marijuana. 

Information on the survey can be found [here](https://news.gallup.com/poll/267698/support-legal-marijuana-steady-past-year.aspx)



## Checking Conditions

$n = 1017$ and $p = 0.66$

1) The survey link indicates that the respondents were chosen from a random sample. We would expect such sample to be independent.   

. . .

2) We need at least 10 people favoring legalizing marijuana and 10 people opposing this. 

$np = 1017 \cdot 0.66 = 671.22$. There are more than 10 people favoring legalizing marijuana.
$n(1-p) = 1017 \cdot (1-0.66) = 345.78$. There are more than 10 people opposing legalizing marijuana.

. . .

##


3) 1017 is less than 10% of US population.


## Confidence Interval 

CI = $\text{point estimate} \pm \text { margin of error}$

. . .

CI = $\text{point estimate} \pm \text { critical value} \times \text{standard error}$

. . .

95% CI for a proportion = $p \pm 1.96 \times \sqrt{\frac{p(1-p)}{n}}$ 

. . .

95% CI = $0.66 \pm 1.96 \times \sqrt{\frac{0.66(1-0.66)}{1017}}$ 

. . .

95% CI = (0.6308857, 0.6891143)

We are 95% confident that the true proportion of Americans who support legalizing marijuana falls between 0.6308857 and 0.6891143.



## Confidence Interval Using R

```{r}
#| echo: true
p <- 0.66 # sample proportion
n <- 1017 # sample size

se <- sqrt(p * (1 - p) / n) # standard error
cv <- qnorm(0.975) # critical value

p - cv * se # lower bound of the CI
p + cv * se # upper bound of the CI
```


# Hypothesis Testing for a Proportion



## Review of Hypothesis Testing

- We always assume the null hypothesis is true at the beginning.

. . .

- We look for evidence against the null.

. . . 

- If we find any evidence against the null then we can conclude null is false. We say we **reject the null hypothesis**.

. . .

##

- If we do not find any evidence against the null then we fail to reject the null. We can keep searching for more evidence against the null. We will never be able to say the null is true so we never accept the null. We **fail to reject the null**.



- We are searching for evidence against the null. 
We are searching for evidence in the sample that may serve as an evidence against the null. 





## Research Question

Do the majority of Americans approve allowing DACA immigrants to become citizens?


## Hypotheses 

$H_0: \pi \leq 0.5$  
$H_A: \pi > 0.5$


# Theoretical reasoning

## Assuming $H_0$ is True

Recall that according to CLT $p \sim \text{approximately }N(\pi, \frac{\pi(1-\pi)}{n})$

. . .

If $H_0: \pi = 0.5$ then the null sampling distribution would be $N(0.5, \frac{0.5(1-0.5)}{n})$


## Looking for Evidence 


According to a Gallup survey of 1520 US adults , 83% approve of allowing DACA immigrants to become citizens.



. . .

$p = 0.83$  
$n = 1520$

. . .

We said that the null sampling distribution would be $N(0.5, \frac{0.5(1-0.5)}{n})$ which is

$N(0.5, \frac{0.5(1-0.5)}{1520})$

$N(0.5, 0.0001644737)$


## The $H_0$ Sampling Distribution 

```{r}
#| echo: false
p <- seq(0.45, 0.55, by = 0.0001)
y <- dnorm(p, mean = 0.5, sd = sqrt(0.5*0.5/1520))
data <- tibble(p = p, y = y)
qplot(p, y, data = data, 
          geom = "line") +
    ylab("") 


```


## What counts as evidence against the null?

Any sample proportion that falls of really far away from the center of the distribution would count as an evidence against the null.

If the null is true, then it would be unlikely to observe extremely high or low sample proportions. 


## Sampling Distribution

```{r}
#| echo: false
se <- sqrt(0.5 * 0.5 / 1520)
p <- seq(0.1, 0.9, by = 0.0001)
y <- dnorm(p, mean = 0.5, sd = se)
data <- tibble(p = p, y = y)
qplot(p, y, data = data, 
          geom = "line") +
  ylab("") +
    geom_vline(xintercept = 0.83) +
   annotate(geom = "text", x = 0.78, y = 20, label = "p = 0.83")


```

##


We want to know the probability of observing an extreme sample proportion like ours (p = 0.83) if the $H_0$ were true.

. . .

If our sample proportion is "extreme" then so is 0.90, 0.91, 0.917, 0.9273423 etc. 



# Hypothesis tests in practice

## 


If the $H_0$ is true what is the probability that we will observe an extremely high sample proportion?


Probability of observing sample proportion of 0.83 and higher 

```{r echo=TRUE}
p <- 0.83
pi_null <- 0.5
n <- 1520
pnorm(p, mean = pi_null, sd = sqrt(pi_null * (1 - pi_null) / n), lower.tail = FALSE)
```

P-value is the probability of observing a sample statistic at least as extreme as the one that has been observed if the null hypothesis were true.

. . .

We can use the cutoff of 0.05 to say the probability is low enough that we reject the null hypothesis. 


## Decision and Conclusion

In the Gallup survey the sample proportion was 0.83. 

. . .

If the null hypothesis ($H_0: \pi = 0.5$) were true, then it would be quite unlikely (p = $2.594241 \times 10^{-146}$) to observe a sample proportion that is at least extreme like the one we observed. 

- We consider this "quite unlikely" since p < 0.05 and reject the $H_0$.

. . .

- We conclude that the population proportion of Americans who approve allowing DACA immigrants to become citizens is higher than 0.5.


## Conditions

`r fontawesome::fa(name = "check", fill = "#e56646")` [Gallup website](https://news.gallup.com/poll/235775/americans-oppose-border-walls-favor-dealing-daca.aspx) indicates that the sample was randomly selected. We will assume independence.

. . .

:::{.pull-left}

`r fontawesome::fa(name = "check", fill = "#e56646")`Number of successes > 10

```{r echo = TRUE}
0.83 * 1520
```

:::

:::{.pull-right}

`r fontawesome::fa(name = "check", fill = "#e56646")`Number of failures > 10

```{r echo = TRUE}
(1 - 0.83) * 1520
```

:::

. . .

<br>
<br>
<br>
<br>

`r fontawesome::fa(name = "check", fill = "#e56646")`1520 < 10% US Population

## Steps of Hypotheses Testing

1. Set hypotheses
2. Identify Sampling Distribution of $H_0$  
3. Calculate p-value  
4. Make a Decision and a Conclusion.

