---
title: "Data 200: Lecture 8a"
subtitle: "Bayesian Credible Intervals and Hypothesis Testing"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

## 

```{r}
library(bayesrules)
library(tidyverse)
```

```{r}
#| echo: false
theme_set(theme_gray(base_size = 18))
```

Examples from this lecture are mainly taken from the [Bayes Rules! book](https://www.bayesrulesbook.com/) and the new functions are from the `bayesrules` package.

## Recall

Last lecture the optimist had the following models.

```{r}
summarize_beta_binomial(alpha = 14, beta = 1, y = 9, n = 20)
```

. . .

Prior model: $\pi \sim \text{Beta}(\alpha = 14, \beta = 1)$

We can read this as the variable $\pi$ follows a Beta model with parameters 14 and 1.

. . .

Posterior model: $\pi|Y \sim \text{Beta}(\alpha = 23, \beta = 12)$

We can read this as $\pi$ given $Y$ (i.e., the data) follows a Beta model with parameters 23 and 12.

## 

```{r}
plot_beta_binomial(alpha = 14, beta = 1, y = 9, n = 20)
```

# Credible Intervals

## 

We are often interested in how our ideas change from prior to posterior. 

One measure that can capture this change is the credible interval. 

## Prior credible interval


```{r}
#| echo: false
ci_prior <- qbeta(c(0.025, 0.975), shape1 = 14, shape2 = 1)
```

According to optimist's prior model the probability that $\pi$ is between 
`r ci_prior[1]` and `r ci_prior[2]` is 95%. 


## 


```{r}
#| echo: false
pi <- seq(0, 1, by = 0.01)
y <- dbeta(pi, shape1 = 14, shape2 = 1)

data <- tibble(pi = pi, y = y)

qplot(pi, y, data = data, geom = "line") +
  geom_ribbon(
    data = subset(data, pi > ci_prior[1] & pi < ci_prior[2]),
    aes(ymax = y),
    ymin = 0,
    fill = "#e56646", 
    colour = NA, 
    alpha = 0.5
  ) +
  annotate(geom = "text", x = 0.9, y = 1.5, label = "95%")
```
##

```{r}
#| echo: false
pi <- seq(0, 1, by = 0.01)
y <- dbeta(pi, 14,1)

data <- tibble(pi = pi, y = y)

qplot(pi, y, data = data, geom = "line") +
  geom_ribbon(
    data = subset(data, pi > ci_prior[1] & pi < ci_prior[2]),
    aes(ymax = y),
    ymin = 0,
    fill = "#e56646",
    colour = NA, 
    alpha = 0.5
  ) +
  annotate(geom = "text", x = 0.9, y = 1.5, label = "95%") +
  geom_segment(
    aes(
      x = ci_prior[1],
      y = dbeta(ci_prior[1], 14, 1),
      xend = 0.65, 
      yend =  5
    ),
    arrow = arrow(length = unit(0.1, "cm"))
  ) +
  annotate(geom = "text", x = 0.65, y = 5.5, label = "2.5th percentile") +
  geom_segment(
    aes(x = 0.99,  y = 12.5, xend = 0.8, yend =  16),
    arrow = arrow(length = unit(0.1, "cm"))
  ) +
  annotate(geom = "text", x = 0.8, y = 17, label = "97.5th percentile")
```
## Prior Credible Interval

```{r}
qbeta(c(0.025, 0.975), shape1 = 14, shape2 = 1)
```

We can utilize the `qbeta()` function to calculate the middle 95% prior credible interval. 

For a given quantile (probability) the `qbeta()` function returns the corresponding $\pi$ value. 

We are essentially calculating the 2.5th and 97.5th percentiles. 

## Posterior credible interval

```{r}
qbeta(c(0.025, 0.975), shape1 = 23, shape2 = 12)
```

```{r}
#| echo: false
posterior_ci <- qbeta(c(0.025, 0.975), shape1 = 23, shape2 = 12)
```

After having observed the data, optimist's posterior model indicates that with 95% probability $\pi$ is between `r posterior_ci[1]` and `r posterior_ci[2]`.

# Hypothesis Testing

## Recall realist's data analysis

```{r}
#| fig-align: center
summarize_beta_binomial(alpha = 5, beta = 11, y = 9, n = 20)

plot_beta_binomial(alpha = 5, beta = 11, y = 9, n = 20) 
```

## Scenario

Let's assume that the general public assumes that more than one-third of the movies pass the Bechdel test. In other words, they believe $\pi \geq 0.33$.

While working on his prior model, the realist was unsure of this and wanted to put this claim to test during his data analysis. 

## Setting hypotheses

$H_0: \pi \geq 0.33$  
$H_A: \pi < 0.33$

The null hypothesis ($H_0$) represents the status quo and the alternative hypothesis, ($H_a$), is realist's claim that he'd like to test.

## Prior Probability

What is the prior probability that $\pi$ is less than 0.33 ? In other words $P(\pi < 0.33) = ?$

. . .


```{r}
prior_prob <- pbeta(0.33, shape1 = 5 , shape2 = 11)
prior_prob
```

Note that `pbeta(0.33, 5, 11)` always shows probability that $pi$ is less than a specific value, in this case 0.33 for a specific Beta distribution, in this case Beta (5,11). 


## Posterior Probability

What is the posterior probability that $\pi$ is less than 0.33 after having observed the data? In other words $P(\pi |Y < 0.33) = ?$

. . .

```{r}
posterior_prob <- pbeta(0.33, shape1 = 14 , shape2 = 22)
posterior_prob
```

##

```{r}
#| echo: false
#| fig-align: center
plot_prob_plot <- function(a,b,cutoff){
  ggplot(data.frame(x = c(0,1)), aes(x=x)) + 
    stat_function(
      fun = dbeta, 
      args = list(a, b), 
      xlim = c(0,cutoff), 
      geom = "area", 
      fill = "lightblue"
    ) + 
    stat_function(fun = dbeta, args = list(a, b)) + 
    geom_segment(
      aes(x = cutoff, xend = cutoff, y = 0, yend = dbeta(cutoff, a, b))
    ) +
    labs(x = expression(pi), y = "density")
}

plot_prob_plot(a = 5, b = 11, cutoff = 0.33) + 
  lims(x = c(0, 1)) +
  labs(title = "Prior")
```

##

```{r}
#| echo: false
#| fig-align: center
plot_prob_plot(a = 14, b = 22, cutoff = 0.33) + 
  lims(x = c(0, 1)) +
  labs(title = "Posterior")
```


## 

$$P(\pi<0.33)$$

```{r}
#| echo: true
prior_prob <- pbeta(0.33, shape1 = 5, shape2 = 11)
prior_prob
```

. . .

```{r}
#| echo: false
prior_odds <- prior_prob / (1 - prior_prob)
```

$$\text{Prior odds } = \frac{P(H_a)}{P(H_0)} = `r prior_odds` \; .$$ 

```{r}
#| echo: true
prior_odds <- prior_prob / (1 - prior_prob)
prior_odds
```

## 

$$P(\pi | Y <0.33)$$

```{r}
#| echo: true
posterior_prob <- pbeta(0.33, shape1 = 14, shape2 = 22)
posterior_prob
```

. . .

```{r}
#| echo: false
posterior_odds <- posterior_prob / (1 - posterior_prob)
```

$$\text{Posterior odds } = \frac{P(H_a |Y)}{P(H_0 |Y)} = `r posterior_odds` \; .$$ 

```{r}
#| echo: true
posterior_odds <- posterior_prob / (1 - posterior_prob)
posterior_odds
```

## 

The __Bayes Factor (BF)__ compares the posterior odds to the prior odds, hence provides insight into just how much the realist's understanding about movies passing the Bechdel test _evolved_ upon observing the sample data:


$$\text{Bayes Factor} = \frac{\text{Posterior odds }}{\text{Prior odds }}$$


```{r}
BF <- posterior_odds / prior_odds
BF
```


## Bayes Factor

In a hypothesis test of two competing hypotheses, $H_a$ vs $H_0$, the Bayes Factor is an odds ratio for $H_a$:

$$\text{Bayes Factor}
= \frac{\text{Posterior odds}}{\text{Prior odds}}
= \frac{P(H_a | Y) / P(H_0 | Y)}{P(H_a) / P(H_0)}
 \; .$$


##

As a ratio, it's meaningful to compare the Bayes Factor (BF)\ to 1.  To this end, consider three possible scenarios:

1. BF = 1:  The plausibility of $H_a$ _didn't change_ in light of the observed data.
2. BF > 1:  The plausibility of $H_a$ _increased_ in light of the observed data.  Thus the greater the Bayes Factor, the more convincing the evidence for $H_a$.
3. BF < 1:  The plausibility of $H_a$ _decreased_ in light of the observed data. 


# Refining hypothesis testing

## Notation

<table align ="center">

<thead>
<th> </th>
<th>Sample Statistic</th>
<th>Population Parameter</th>
</thead>

<tr> 
<td>Proportion</td>
<td>p</td>
<td>&pi;</td>

</tr>


</table>

## Notation

<table align ="center">

<thead>
<th> </th>
<th>Sample Statistic</th>
<th>Population Parameter</th>
</thead>

<tr> 
<td>Proportion</td>
<td>p</td>
<td>&pi;</td>

</tr>

<tr>
<td>Mean</td>
<td> x&#772; </td>
<td> &mu;</td>

</tr>

<tr>
<td> Standard Deviation </td>
<td> s </td>
<td> &sigma;</td>

</tr>

<tr>
<td> Variance </td>
<td> s<sup>2</sup> </td>
<td> &sigma;<sup>2</sup></td>

</tr>

</table>

*There are more types of hypothesis tests beyond what we cover in this class, but they all follow the same structure!*

##

To make the connection between research questions of studies and statistics, we will take small steps and begin with writing hypotheses using notation.

##

`r fontawesome::fa(name = "search", fill = "#e56646")` **Research Question** Do the majority of Americans approve allowing DACA immigrants to become citizens?

. . .

`r fontawesome::fa(name = "table", fill = "#e56646")` **Variable** `approve` (yes, yes, yes, no, yes, no, no)

. . .

`r fontawesome::fa(name = "language", fill = "#e56646")` **Research Question Using Notation** $\pi \stackrel{?}{>} 0.5$ 

. . .

**Hypotheses** 

$H_0: \pi \leq 0.5$  
$H_A: \pi > 0.5$

. . .

The parameter we want to infer about is **a single proportion**.

##

::: {.callout-tip }

If you want to type math notation correctly in Quarto as $\pi$ then you can write 

```{r}
#| echo: true
#| eval: false
$$\pi$$
```

The double dollar signs at the beginning and at the end let Quarto know that you are writing a math equation. You would do this in regular text, not a code chunk.

:::

##

`r fontawesome::fa(name = "search", fill = "#e56646")` **Research Question** Is California March 2020 unemployment rate different than US March 2020 unemployment rate which is at 4.4%?

. . .

`r fontawesome::fa(name = "table", fill = "#e56646")` **Variable** `unemployed_CA` (no, no, yes, no, yes, no, no...)

. . .

`r fontawesome::fa(name = "language", fill = "#e56646")` **Research Question Using Notation** $\pi \stackrel{?}{=} 0.044$

. . .

**Hypotheses**

$$H_0:\pi= 0.044$$  $$H_A: \pi \neq 0.044$$

. . .


The parameter we want to infer about is **a single proportion**.

## 

`r fontawesome::fa(name = "search", fill = "#e56646")` **Research Question** Are there more STEM majors at CS UCI than non-STEM majors?

. . .

`r fontawesome::fa(name = "table", fill = "#e56646")` **Variable** `STEM` (TRUE, FALSE, FALSE, TRUE, FALSE, TRUE, TRUE...)

. . .

`r fontawesome::fa(name = "language", fill = "#e56646")` **Research Question Using Notation** $\pi_{STEM} \stackrel{?}{>} 0.5$

. . .

**Hypotheses**

$$H_0: \pi \leq 0.5$$ $$H_A: \pi > 0.5$$

. . .

The parameter we want to infer about is **a single proportion**.

##

`r fontawesome::fa(name = "search", fill = "#e56646")` **Research Question** Do CS UCI students sleep on average 8 hours on a typical night? 


. . .

`r fontawesome::fa(name = "table", fill = "#e56646")` **Variable** `sleep` (8,7,9,7.5, ...)

. . .

`r fontawesome::fa(name = "language", fill = "#e56646")` **Research Question Using Notation** $\mu \stackrel{?}{=} 8$

. . .

**Hypotheses** 

:::{.pull-left}
$H_0 : \mu = 8$  
$H_A : \mu \neq 8$

::: 

:::{.pull-right}
$H_0 : \mu - 8 = 0$  
$H_A : \mu - 8 \neq 0$ 
:::

. . .

The parameter we want to infer about is **a single mean**.

##

`r fontawesome::fa(name = "search", fill = "#e56646")` **RQ** Do Democrats and Republicans approve legal abortion at same rates?

. . .

`r fontawesome::fa(name = "table", fill = "#e56646")` **Variables** explanatory: `party` (D, D, R, R,...)  
response: `approve`(TRUE, FALSE, FALSE, TRUE,...)

. . .

`r fontawesome::fa(name = "language", fill = "#e56646")` **Research Question Using Notation** $\pi_{d} \stackrel{?}{=} \pi_{r}$ or $\pi_{d} -  \pi_{r} \stackrel{?}{=}0$ 

. . .

**Hypotheses**

$H_0:\pi_{d} = \pi_{r}$  
$H_A:\pi_{d} \neq \pi_{r}$
. . .

We want to infer about **difference of two proportions**.


##


<div align="center">

|                               | Parameter of Interest | Response variable             |Explanatory variable           |
|-------------------------------|-----------------------|-----------------------|-----------------------|
| Single Proportion             | $\pi$                 | Binary                |
| Difference of Two Proportions | $\pi_1 - \pi_2$       | Binary                |Binary                 |
| Single Mean                   | $\mu$                 | Numeric               |
| Difference of Two Means       | $\mu_1 - \mu_2$       | Numeric               |Binary                 |

</div>

A categorical variable with two levels is called a __binary__ variable.

##

Later on we will also learn

<div align = "center">

| Parameter of Interest | Response              |Explanatory                |
|-----------------------|-----------------------|-----------------------    |
| $\beta_1$             | Numeric               | Categorical and/or Numeric|

</div>
