---
title: "Data 200: Lecture 11A"
subtitle: "Multiple Linear Regression, categorical predictors, transformations, model evaluation"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
options(scipen = 0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)
theme_set(theme_bw(base_size = 22))
```

## 

Last week we learned about simple linear regression

$$y_i = \beta_0 +\beta_1x_{1i} + \epsilon_i$$

Last class we learned how to check if the conditions for simple linear regression were met.

Today we will learn how we can adjust our model.

##

In your homework you looked at the `penguins` data.

Let's see if we can model bill depth from flipper length

<center>

| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Bill depth | $y$ | Response | Numeric |
| Flipper length | $x_1$ | Explanatory | Numeric |

</center>

##

```{r}
ggplot(penguins, aes(x = flipper_len, y = bill_dep)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Model checking

- Independence (need context)

- Linearity (plot residual)

- Constant Variance (plot residuals)

- Normality of Residuals (plot residuals)

## Model checking: Linearity

Do the residuals look random or do you see a pattern?

```{r}
penguins_model_1 <- lm(bill_dep ~ flipper_len, data = penguins)

fortify(penguins_model_1) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

. . .

This pattern could be due to an unmodeled variable.


##

**Simple** linear regression = only one $X$

$$y_i = \beta_0 +\beta_1x_{1i} + \epsilon_i$$

**Multiple** linear regression = multiple $X$'s

$$y_i = \beta_0 +\beta_1x_{1i} + \beta_1 x_{2i} + ... + \beta_p x_{pi} \epsilon_i$$

##

```{r}
ggplot(penguins, aes(x = flipper_len, y = bill_dep, color = species)) + 
  geom_point()
```

##

Let's also consider species type in our model

<center>

| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Bill depth | $y$ | Response | Numeric |
| Flipper length | $x_1$ | Explanatory | Numeric |
| Species | $x_2$ | Explanatory | Categorical |

</center>

## Model checking: Linearity

Do the residuals look random or do you see a pattern?

```{r}
penguins_model_2 <- lm(bill_dep ~ flipper_len + species, data = penguins)

fortify(penguins_model_2) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

. . .

*The residuals are looking more random now.*

## Comparing models

```{r}
tidy(penguins_model_1)
```

Predicted Bill Depth = `r round(penguins_model_1$coefficients[1], digits = 1)` + `r round(penguins_model_1$coefficients[2], digits = 1)` * Flipper Length

. . .

```{r}
tidy(penguins_model_2)
```

Predicted Bill Depth = `r round(penguins_model_2$coefficients[1], digits = 1)` + `r round(penguins_model_2$coefficients["flipper_len"], digits = 1)` * Flipper Length + `r round(penguins_model_2$coefficients["speciesChinstrap"], digits = 1)` I(Species = Chinstrap) + `r round(penguins_model_2$coefficients["speciesGentoo"], digits = 1)` I(Species = Gentoo)

# Categorical predictors

## Categorical $X$'s

**Predicted Bill Depth = `r round(penguins_model_2$coefficients[1], digits = 1)` + `r round(penguins_model_2$coefficients["flipper_len"], digits = 1)` * Flipper Length + `r round(penguins_model_2$coefficients["speciesChinstrap"], digits = 1)` I(Species = Chinstrap) + `r round(penguins_model_2$coefficients["speciesGentoo"], digits = 1)` I(Species = Gentoo)**

The I() function only gives value 1 if true, and 0 if false

- These are used for categorical predictors
- Recall there are three levels for species, but notice only 2 species levels appear in our model 
- **Reference level** is the omitted level for a categorical variable 
- For every categorical variable with $k$ levels, there will be $k-1$ predictors in the model (because of math...)

##

**Predicted Bill Depth = `r round(penguins_model_2$coefficients[1], digits = 1)` + `r round(penguins_model_2$coefficients["flipper_len"], digits = 1)` * Flipper Length + `r round(penguins_model_2$coefficients["speciesChinstrap"], digits = 1)` I(Species = Chinstrap) + `r round(penguins_model_2$coefficients["speciesGentoo"], digits = 1)` I(Species = Gentoo)**

Predicted Bill Depth for a Gentoo penguin with a flipper length of 200mm:

```{r}
#| echo: false

pred_1 <- penguins_model_2$coefficients[1] + penguins_model_2$coefficients["flipper_len"] * 200 + penguins_model_2$coefficients["speciesChinstrap"] * 0 + penguins_model_2$coefficients["speciesGentoo"] * 1
```


`r round(penguins_model_2$coefficients[1], digits = 1)` + `r round(penguins_model_2$coefficients["flipper_len"], digits = 1)` * 200 + `r round(penguins_model_2$coefficients["speciesChinstrap"], digits = 1)` x 0 + `r round(penguins_model_2$coefficients["speciesGentoo"], digits = 1)` x 1 = `r round(pred_1, digits = 1)`

. . .

Predicted Bill Depth for a Chinstrap penguin with a flipper length of 200mm:

```{r}
#| echo: false

pred_2 <- penguins_model_2$coefficients[1] + penguins_model_2$coefficients["flipper_len"] * 200 + penguins_model_2$coefficients["speciesChinstrap"] * 1 + penguins_model_2$coefficients["speciesGentoo"] * 0
```

`r round(penguins_model_2$coefficients[1], digits = 1)` + `r round(penguins_model_2$coefficients["flipper_len"], digits = 1)` * 200 + `r round(penguins_model_2$coefficients["speciesChinstrap"], digits = 1)` x 1 + `r round(penguins_model_2$coefficients["speciesGentoo"], digits = 1)` x 0 = `r round(pred_2, digits = 1)`

. . .

Predicted Bill Depth for an Adelie penguin with a flipper length of 200mm:

```{r}
#| echo: false

pred_3 <- penguins_model_2$coefficients[1] + penguins_model_2$coefficients["flipper_len"] * 200 + penguins_model_2$coefficients["speciesChinstrap"] * 0 + penguins_model_2$coefficients["speciesGentoo"] * 0
```

`r round(penguins_model_2$coefficients[1], digits = 1)` + `r round(penguins_model_2$coefficients["flipper_len"], digits = 1)` * 200 + `r round(penguins_model_2$coefficients["speciesChinstrap"], digits = 1)` x 0 + `r round(penguins_model_2$coefficients["speciesGentoo"], digits = 1)` x 0 = `r round(pred_3, digits = 1)`

##

Let's also consider body mass type in our model

<center>

| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Bill depth | $y$ | Response | Numeric |
| Flipper length | $x_1$ | Explanatory | Numeric |
| Species | $x_2$ | Explanatory | Categorical |
| Body mass | $x_3$ | Explanatory | Numeric |

</center>


## 

It gets harder/impossible to visualize our model the more predictors we have. Instead we can still look at a plot of $Y$ and one $X$ to see if we think they are linearly related.

```{r}
ggplot(penguins, aes(x = flipper_len, y = body_mass)) + 
  geom_point()
```


## Comparing models

```{r}
penguins_model_2 <- lm(bill_dep ~ flipper_len + species, data = penguins)
tidy(penguins_model_2)
```


```{r}
penguins_model_3 <- lm(bill_dep ~ flipper_len + species + body_mass, data = penguins)
tidy(penguins_model_3)
```


## Model checking: Constant variance of residuals

Do the residuals appear to have constant variance?

```{r}
fortify(penguins_model_3) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

. . .

*There does seem to be higher variance for higher fitted/predicted values. A possible solution could be to transform the response variable.*

# Transformations

Common transformations to correct for non constant variance of residuals:

- $log(Y)$, $log(X)$
- $\sqrt{Y}, \sqrt{X})$

##

Let's consider log(bill depth) as our response variable

::::{.columns}

:::{.column width="50%"}

```{r}
ggplot(
  penguins, 
  aes(
    x = flipper_len, 
    y = bill_dep, 
    color = species
  )
) +
  geom_point() 
```

:::

:::{.column width="50%"}

```{r}
ggplot(
  penguins, 
  aes(
    x = flipper_len, 
    y = log(bill_dep), 
    color = species
  )
) +
  geom_point()
```

:::

::::

##

Let's also consider body mass type in our model

<center>

| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| log(Bill depth) | $y$ | Response | Numeric |
| Flipper length | $x_1$ | Explanatory | Numeric |
| Species | $x_2$ | Explanatory | Categorical |
| Body mass | $x_3$ | Explanatory | Numeric |

</center>

. . .

Note in `R` log() is natural log() = \ln() = $\text{log}_e$()

## 


```{r}
penguins_model_4 <- lm(log(bill_dep) ~ flipper_len + species + body_mass, data = penguins)

tidy(penguins_model_4)
```

**Predicted log(Bill Depth)** = `r round(penguins_model_4$coefficients[1], digits = 1)` + `r round(penguins_model_4$coefficients["flipper_len"], digits = 1)` Flipper Length + `r round(penguins_model_4$coefficients["speciesChinstrap"], digits = 1)` I(Species = Chinstrap) + `r round(penguins_model_4$coefficients["speciesGentoo"], digits = 1)` I(Species = Gentoo) + `r round(penguins_model_4$coefficients["body_mass"], digits = 1)` Body Mass

. . .

**Predicted Bill Depth** = e\^\{`r round(penguins_model_4$coefficients[1], digits = 1)` + `r round(penguins_model_4$coefficients["flipper_len"], digits = 1)` Flipper Length + `r round(penguins_model_4$coefficients["speciesChinstrap"], digits = 1)` I(Species = Chinstrap) + `r round(penguins_model_4$coefficients["speciesGentoo"], digits = 1)` I(Species = Gentoo) + `r round(penguins_model_4$coefficients["body_mass"], digits = 1)` Body Mass \}

## Model checking: Constant variance of residuals

Do the residuals appear to have constant variance?

```{r}
fortify(penguins_model_4) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

. . .

Did we make a better model?


# Model Evaluation

## Common evaluation metrics

- $R^2$
- Adjusted $R^2$
- Root Mean Square Error (RMSE)

*For models that meet the conditions of linear regression.*

## $R^2$

$R^2$ is the percent of variation on $Y$ explained by $X$

- Informs about how well the model fits the data
- Bounded between 0 and 1

## $R^2$

```{r}
glance(penguins_model_3)
```

```{r}
glance(penguins_model_4)
```

Model 4 explains 83% of the variation in log(Bill Depth).

Model 3 explains 81% of the variation in Bill Depth.

So model 4 appears to be a slightly better fit for the data.

## Adjusted $R^2$

A major issue with $R^2$ is that it will always increase when you add another predictor $X$, even if it is not a better model.

A more responsible metric to use is Adjusted $R^2$, which account for the number of predictors in your model

## Adjusted R-Squared

When comparing models with multiple predictors we rely on Adjusted R-squared.


```{r}
glance(penguins_model_2)
```

```{r}
glance(penguins_model_4)
```

Adjusted $R^2$ supports that model 4 (Adj $R^2$ = `r round(glance(penguins_model_4)$adj.r.squared, digits = 2)`) is a better fit for our data than model 2 (Adj $R^2$ = `r round(glance(penguins_model_2)$adj.r.squared, digits = 2)`).


## Root Mean Square Error

$RMSE = \sqrt{\frac{\Sigma_{i=1}^n(y_i-\hat y_i)^2}{n}}$

- "Essentially" measuring average prediction error for our model
- Lowest possible value is 0
- Same units as $Y$

##

```{r}
rmse(penguins_model_2, data = penguins)
```

The predictions from model 2 are, on average, off by 0.97 mm. 

```{r}
rmse(penguins_model_3, data = penguins)
```

The predictions from model 3 are, on average, off by 0.86 mm. 

## Units of RMSE

```{r}
rmse(penguins_model_3, data = penguins)
```

The predictions for bill depth from model 3 are, on average, off by 0.86 mm.

```{r}
rmse(penguins_model_4, data = penguins)
```

The predictions for log(bill depth) from model 4 are, on average, off by 0.05 mm.

We should not use RMSE to compare models with different $y$'s.


# Overfitting

Can we keep adding all the variables and try to get an EXCELLENT model fit?

## Overfitting

- We are fitting the model to sample data.

- Our goal is to understand the population data.

- If we make our model too perfect for our sample data, the model may not perform as well with other sample data from the population.

- In this case we would be overfitting the data.

- We can use **model validation** techniques.

## Splitting the data

To avoid overfitting we can split our data

- Some of the data will be reserved for fitting/training our model
- The rest of the data will be reserved for evaluating the prediction of our model

## Splitting the data (Train vs. Test)

```{r message=FALSE}
set.seed(8236)
library(rsample)

penguins_split <- initial_split(penguins, prop = 0.75) ## 75% to 25% split
```

. . .

```{r}
dim(penguins)
```

. . .

```{r}
penguins_train_data <- training(penguins_split)
dim(penguins_train_data)
```

. . .

```{r}
penguins_test_data <- testing(penguins_split)
dim(penguins_test_data)
```

##

Fit the model with the training data
```{r}
model_3_train <- lm(
  bill_dep ~ flipper_len + species + body_mass,  
  data = penguins_train_data
)
```

Predict bill length for penguins in testing data
```{r}
model_3_test_preditions <- predict(model_3_train, newdata = penguins_test_data)
```

##

In sample (train) RMSE
```{r}
modelr::rmse(
  model_3_train, 
  data = penguins_train_data
)
```

Our of sample (test) RMSE
```{r}
predicted_y <- model_3_test_preditions
actual_y <- penguins_test_data$bill_dep

sqrt(mean((predicted_y - actual_y)^2, na.rm = TRUE))
```

Does it make sense that our in-sample RMSE > out-of-sample RMSE?

