---
title: "Data 200: Lab 12 Solution"
subtitle: "Logistic predictive modeling competion"
author: "Your name here"
format: pdf
---

All of the packages listed below have been preinstalled, so please do not install packages again. You can just run the r chunk below to load all of the packages.
```{r}
library(broom)
library(rsample)
library(openintro)
library(modelr)
library(caret)
library(tidyverse)

data("email")
```


This week we are going to build predictive logistic models to detect if emails are spam, using the `email` dataset.

This means you will:

1. Split the email data into a training (65% of your sample) and a testing set (35% of your sample). Before you do this make sure to set your seed to 20251112.
2. Build a model. It is up to you to choose which variables available in the data you are going to include in your model.
3. Asses out-of-sample predictive error, accuracy, on the test data set. I was able to get an out-of-sample predictive accuracy of 0.9104, can you beat that?
4. Submit your lab 12 on Canvas/CI Learn by the end of the day for extra credit.


**Solution**

Split data
```{r}
set.seed(20251112)

email_split <- initial_split(email, prop = 0.65) 
email_train_data <- training(email_split)
email_test_data <- testing(email_split)
```

Fit the logistic model. This is where it is up to you to choose which predictors to incorporate.
```{r}
spam_model <- glm(
  spam ~ to_multiple + winner + format +
    re_subj + exclaim_subj +
    cc + attach + dollar +
    inherit + password,
  data = email_train_data,
  family = binomial
)

```

Compute confusion matrix
```{r}
email_test_data <- email_test_data |> 
  add_predictions(spam_model) |> 
  mutate(pred_p = exp(pred)/(1 + exp(pred))) |> 
  mutate(pred_y = case_when(pred_p < 0.5 ~ 0, pred_p >= 0.5 ~ 1))

confusionMatrix(
  as.factor(email_test_data$pred_y),
  as.factor(email_test_data$spam)
)
```

The out-of-sample accuracy of my model is 0.9104.


