---
title: "Data 200: Lecture 12A"
subtitle: "Logistic regression"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
options(scipen = 0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)
theme_set(theme_bw(base_size = 22))
```


## (Normal) Linear Regression Response Variables

- Birth weight of Babies (55 - 176 ounces) 
- Sale Prices ($12789 - $755,000)
- Number of Species (6 - 129 mammals)


## Logistic Regression Response Variables

- Will it rain tomorrow? (Yes/No) 
- Is email spam? (Yes/No)
- Does the candidate receive a callback? (Yes/No)

. . .

When the response variable is binary a logistic regression model can be utilized.

##

[Are Emily and Greg More Employable than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination](https://www.nber.org/papers/w9873). 


- Researchers respond to help-wanted ads in Boston and Chicago newspapers with fictitious resumes.

. . .

- They randomly assign White sounding names to half the resumes and African American sounding names to the other half.

##

[Are Emily and Greg More Employable than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination](https://www.nber.org/papers/w9873).

- They create high quality resumes (more experience, likely to have an email address etc.) and low quality resumes. 

- For each job ad they send four resumes (two high quality and two low quality.)

## Data

```{r echo=FALSE, message=FALSE}
data(resume)
resume <- resume |> 
  mutate(received_callback = as.logical(received_callback))
```

```{r}
resume <- resume |> 
  select(received_callback, race, years_experience, job_city)

glimpse(resume)
```


##

Response variable: `received_callback`

:::{.pull-left}

```{r}
count(resume, received_callback) |> 
  mutate(prop = n / sum(n))
```

:::

:::{.pull-right}
```{r fig.height=6, echo = FALSE}
ggplot(resume, aes(x = received_callback)) +
  geom_bar()
```

:::


## Notation

$y_i$ = whether a (fictitious) job candidate receives a call back.

$\pi_i$ = probability that the $i$th job candidate will receive a call back.

$1-\pi_i$ = probability that the $i$th job candidate will **not** receive a call back.


## Where is the line?

```{r fig.align='center', fig.height=4}
ggplot(resume, aes(x = race, y = received_callback)) +
  geom_point()
```


## The Linear Model

We can model the probability of receiving a callback with a linear model.

$\text{transformation}(\pi_i) = \beta_0 + \beta_1x_{1i}+\beta_2x_{2i} +.... \beta_kx_{ki}$


. . .

$logit(\pi_i) = \beta_0 + \beta_1x_{1i}+\beta_2x_{2i} +.... \beta_kx_{ki}$

. . .

$logit(\pi_i) = log(\frac{\pi_i}{1-\pi_i})$


##

**Probability** $\pi_i$ Probability of receiving a callback.

. . .

**Odds** $\frac{\pi_i}{1-\pi_i}$ Odds of receiving a callback.

. . .

**Logit** $log(\frac{\pi_i}{1-\pi_i})$ Logit of receiving a callback.

## Probability and Odds

What are odds?
<!-- $$Odds = \frac{\text{Probability of success}}{\text{Probability of failure}} = \frac{P(Y = 1)}{P(Y = 0)} = \frac{P(Y=1)}{1 - P(Y=1)} $$ -->

Relationship between probability and odds
<!-- $$\text{Probability} = \frac{\text{Odds}}{1 + \text{Odds}}$$ -->


```{r}
#| fig-height: 3.75
#| echo: false

prob_seq <- seq(0, 1, by = 0.01)

prob_odds_df <- data.frame(
  "probability" = prob_seq,
  "odds" = prob_seq / (1 - prob_seq),
  "log_odds" = log(prob_seq / (1 - prob_seq))
)

gridExtra::grid.arrange(
  ggplot(prob_odds_df, aes(x = probability, y = odds)) +
    geom_point() +
    labs(x = "Probability", y = "Odds") +
    theme_bw(base_size = 20),
  ggplot(prob_odds_df, aes(x = probability, y = log_odds)) +
    geom_point() +
    labs(x = "Probability", y = "Log Odds / Logit") +
    theme_bw(base_size = 20),
  nrow = 1
)
```

##

| Scale       | Range                  |
|-------------|------------------------|
| Probability | 0 to 1                 |
| Odds        | 0 to $\infty$          |
| Logit       | - $\infty$ to $\infty$ |

##

```{r fig.height=6, echo=FALSE, fig.align='center'}
ggplot(resume, aes(x = race, fill = received_callback)) +
  geom_bar(position = "fill") +
  ylab("proportion")
```



## When race is black

```{r}
resume |> 
  filter(race == "black") |> 
  count(received_callback) |> 
  mutate(prop = n / sum(n))
```



## When race is black

```{r}
resume |> 
  filter(race == "black") |> 
  count(received_callback) |> 
  mutate(prop = n / sum(n))
```

```{r}
#| echo: false

p_b <- resume |> 
  filter(race == "black") |> 
  count(received_callback) |> 
  mutate(prop = n / sum(n)) |> 
  filter(received_callback == TRUE) |> 
  select(prop) |> 
  pull()
  
```

Probability of receiving a callback when the candidate has a Black sounding name is `r p_b`.


## When race is white

```{r}
resume |> 
  filter(race == "white") |> 
  count(received_callback) |> 
  mutate(prop = n / sum(n))
```

```{r}
#| echo: false

p_w <- resume |> 
  filter(race == "white") |> 
  count(received_callback) |> 
  mutate(prop = n / sum(n)) |> 
  filter(received_callback == TRUE) |> 
  select(prop) |> 
  pull()
```

Probability of receiving a callback when the candidate has a white sounding name is `r p_w`.

##

:::{.pull-left}


```{r}
p_b
## Odds
odds_b <- p_b / (1 - p_b)
odds_b

## Logit
logit_b <- log(odds_b)
logit_b
```


:::

. . .

:::{.pull-right}

```{r}
p_w
## Odds
odds_w <- p_w / (1 - p_w)
odds_w

## Logit
logit_w <- log(odds_w)
logit_w
```

:::

## Linear model for log odds of call back

R converts `black` to 0 and `white` to 1 using alphabetical ordering

```{r}
#| echo: false
result_data <- tibble(
  race = c(0, 1),
  logit = c(logit_b, logit_w)
)

coords <- paste(result_data$race, round(result_data$logit,2), sep = ", ")

logit_plot <- ggplot(result_data, aes(x = race, y = logit)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = c(0, 1)) +
  geom_label(aes(race + .1, logit, label = coords)) +
  xlim(-0.05,1.2)
```

```{r}
#| fig-align: center
#| fig-height: 4
#| echo: false
logit_plot
```

##

```{r}
#| fig-align: center
#| fig-height: 3
#| echo: false
logit_plot
```

The intercept of the line is 

```{r}
logit_b
```

##

```{r}
#| fig-align: center
#| fig-height: 3
#| echo: false
logit_plot
```

The slope of the line is:

```{r}
(logit_w - logit_b) / 1 # rise / run
```

As x increases by 1 unit, the expected change in the logit of receiving call back is `r logit_w - logit_b`. In this case, this is just the difference between logit for the white group and the black group. 

##

```{r}
model_r <- glm(
  received_callback ~ race, 
  data = resume, 
  family = binomial
)
```

```{r}
tidy(model_r)
```

$log(\frac{\hat \pi_i}{1-\hat \pi_i}) = -2.67 + 0.438\times I(race_i = white)$

## Pause

. . .

It can be hard to understand the "logit / log odds", but the main thing to remember is: 

increase in probability = increase in odds = increase in log odds.

. . .

We built a model for probability of a binary random variable.

- We use the `glm()` function with `family = binomial` to fit the logistic regression model
- It will predict log odds, but we can always convert back to probability

##

We will consider years of experience as an explanatory variable. Normally, we would also include race in the model and have multiple explanatory variables, however, for learning purposes, we will keep the model simple.

```{r}
model_y <- glm(
  received_callback ~ years_experience, 
  data = resume, 
  family = binomial
)
```

```{r}
tidy(model_y)
```

## Extracting estimates for reporting with inline R code

```{r}
model_y_summary <- tidy(model_y)

intercept <- model_y_summary |> 
  filter(term == "(Intercept)") |> 
  select(estimate) |> 
  pull()

slope <- model_y_summary |> 
  filter(term == "years_experience") |> 
  select(estimate) |> 
  pull()
```

Logit of receiving a callback for a candidate with $x$ years of experience is

logit callback = log(odds call back) = `r round(intercept, digits = 2)` + `r round(slope, digits = 4)` x



## From logit to odds

Model: logit callback = `r round(intercept, digits = 2)` + `r round(slope, digits = 4)` Years Experience

Logit for a Candidate with 1 year of experience 

$-2.76 +  0.0391 \times 1$

. . .

Odds for a Candidate with 1 year of experience

$e^{logit} = odds = \frac{\pi_i}{1 - \pi_i}$

so

Odds $= \frac{\hat\pi_i}{1-\hat\pi_i} = e^{-2.76 + 0.0391 \times 1}$



## From odds to probability

$\pi_i = \frac{odds}{1+odds}$

$\hat\pi_i = \frac{e^{-2.76 + 0.0391 \times 1}}{1+e^{-2.76 + 0.0391 \times 1}} = 0.0618$

```{r}
exp(-2.76 + 0.0391 * 1) / (1 + exp(-2.76 + 0.0391 * 1))
```

. . .

Note you can use `exp()` function in R for exponentiating number e.

```{r}
exp(1)
```


## Logistic Regression model 

**Logit form:** 

$log(\frac{\pi_i}{1-\pi_i}) = \beta_0 + \beta_1x_{1i}+\beta_2x_{2i} +.... \beta_kx_{ki}$

**Probability form:**

$\large{\pi_i = \frac{e^{\beta_0 + \beta_1x_{1i}+\beta_2x_{2i} +.... \beta_kx_{ki}}}{1+e^{\beta_0 + \beta_1x_{1i}+\beta_2x_{2i} +.... \beta_kx_{ki}}}}$




:::{.pull-left}

Estimated probability of a candidate with 0 years of experience receiving a callback

$\hat\pi_i = \frac{e^{-2.76 + 0.0391 \times 0}}{1+e^{-2.76 + 0.0391 \times 0}} = 0.0595$

:::

:::{.pull-right}

Estimated probability of a candidate with 1 year of experience receiving a callback

$\hat\pi_i = \frac{e^{-2.76 + 0.0391 \times 1}}{1+e^{-2.76 + 0.0391 \times 1}} = 0.0618$

:::

##


```{r}
model_ryc <- glm(
  received_callback ~ race + years_experience + job_city,
  data = resume,
  family = binomial
)
```

##


```{r}
tidy(model_ryc)
```

##

The estimated probability that a Black candidate with 10 years of experience, residing in Boston, would receive a callback.

$\large{\hat\pi_i = \frac{e^{-2.78 + (0.0440 \times 0) + (0.0332\times10) + (-0.0329\times 0)}}{1+e^{-2.78 + (0.0440 \times 0) + (0.0332\times10) + (-0.0329\times 0)}} = 0.0796}$

```{r}
large_exponent <- -2.78 + (0.0440 * 0) + (0.0332 * 10) + (-0.0329 * 0)

exp(large_exponent) / (1 + exp(large_exponent))
```


##

We have used the data for educational purposes. 
The original study considers many other variables that may influence whether someone receives a callback or not. Read the original study for other considerations.

[Are Emily and Greg More Employable than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination](https://www.nber.org/papers/w9873). 

# Model Evaluation

How do we know if our logitistic regression model is predicting well?

## Model evaluation example

Let's model probability of low baby birth weight on gestation

```{r}
# Creating binary variable
babies <- babies |> 
  mutate(low_bwt = case_when(bwt < 88 ~ TRUE, bwt >= 88~ FALSE)) |> 
  drop_na(gestation)
```

`low_bwt` is a binary variable, either `TRUE` or `FALSE`

. . .

```{r}
# Fitting logistic regression model
model_g <- glm(low_bwt ~ gestation, data = babies, family = "binomial")
tidy(model_g)
```



##

Remember that our model will predict logit of callback, but our response variable is a binary (yes or no).

How do we decide if our model is predicting well?

. . .

- we can convert predicted logit of callback to probability of call back
- then use a cutoff to convert predicted probability of callback to yes/no. For example:
  - If predicted probability callback $\geq$ 0.5
    - predict `yes` callback
  - If predicted probability callback < 0.5
    - predict `no` callback

##

```{r}
model_g <- glm(low_bwt ~ gestation, data = babies, family = "binomial")

tidy(model_g)
```

$\hat{\pi} = \frac{\exp(b_0 +b_1x)}{1+\exp(b_0 + b_1x)}$

$\hat{\pi}$ when gestation is 284 oz = $\frac{\exp(17.5 -0.0758 \cdot 284)}{1+\exp(17.5 -0.0758 \cdot 284)} = \frac{\exp(-4.0272)}{1+ \exp(-4.0272)} = 0.01751203$

. . .

*We won't actually do this ourselves, we can use a function!*

##

```{r}
babies <- babies |> 
  add_predictions(model_g) 

select(babies, low_bwt, pred) 
```

##

```{r}
babies <- babies |>
  add_predictions(model_g) |> 
  mutate(pred_p = exp(pred)/(1 + exp(pred)))

select(babies, low_bwt, pred, pred_p)
```

## Cutoff 

```{r}
babies <- babies |> 
  add_predictions(model_g) |> 
  mutate(pred_p = exp(pred)/(1 + exp(pred))) |> 
  mutate(pred_y = case_when(pred_p < 0.5 ~ FALSE, pred_p >= 0.5 ~ TRUE))

select(babies, low_bwt, pred, pred_p, pred_y)
```

## Confusion Matrix

```{r}
caret::confusionMatrix(
  as.factor(babies$pred_y), # Predicted response
  as.factor(babies$low_bwt) # Actual response (from our data)
)
```

**Accuracy**: (1161 + 4) / (1161 + 4 + 53 + 4) = 0.9526

# Model validation

Recall last week we learned about splitting our data to avoid over fitting to the sample

- Part of the data (training data) is used to fit the model
- the rest to of the data (testing data) is used assess out-of-sample prediction error. 

We can do the same here

## Split data

```{r}
set.seed(8236)
library(rsample)

babies_split <- initial_split(babies, prop = 0.60) ## 60% to 40% split
babies_train_data <- training(babies_split)
babies_test_data <- testing(babies_split)
```

## Fit model and convert predictions

```{r}
model_g <- glm(low_bwt ~ gestation, data = babies_train_data, family = "binomial")

babies_test_data <- babies_test_data |> 
  add_predictions(model_g) |> 
  mutate(pred_p = exp(pred)/(1 + exp(pred))) |> 
  mutate(pred_y = case_when(pred_p < 0.5 ~ FALSE, pred_p >= 0.5 ~ TRUE))
```

## Out-of-sample prediction accuracy

```{r}
caret::confusionMatrix(
  as.factor(babies_test_data$pred_y), # Predicted response
  as.factor(babies_test_data$low_bwt) # Actual response (from our data)
)
```

