---
title: "Data 200: Lab 3a Solutions"
subtitle: "Plotting foundations"
author: "Professor Medina"
format: html
---

*Please always start labs and homework by changing the author to your name. Please answer questions using complete sentences where appropriate.*

## Organizing our knowledge: Appropriate plot by variable type

### Question 1

Below are some plot types we discussed in class.
When you want to plot data you first have to think about what types of plot are appropriate.

Below are the plot types we discussed in class. 
Write each under the data scenario they are appropriate for.
The first is done for you as an example.

#### Plots discussed

1. bar plot
2. standardized bar plot
3. dodged bar plot
4. histogram
5. side-by-side histograms
6. box plot
7. side-by-side box plots
8. scatter plot
9. scatter plot with points differentially colored

#### Data scenarios

**Solution:**

Single categorical variable:

- bar plot

Single numeric variable:

- box plot
- histogram

Two categorical variables:

- standardized bar plot
- dodged bar plot

Two numeric variables:

- scatter plot

One numeric variable for different levels of a categorical variable:

- side-by-side histograms
- side-by-side box plots

Two numeric variables for different levels of a categorical variable:

- scatter plot with points differentially colored


## Question 2

```{r}
library(openintro)
library(tidyverse)

## We will use the `movies` data from the `openintro` package
glimpse(movies)
```

Look at the help documentation for the `movies` data. What year were these movies released in?

**Solution:**

```{r}
#| eval: false
?movies
```

## Question 3

Produce a plot to visualize the relationship between critic score and box office earning.

**Solution:**

```{r}
ggplot(movies, aes(x = score, y = box_office)) +
  geom_point()
```

## Question 4

Do you see any relationship between critic score and box office earning? Another way to think about this is "If I had information about what the critic score for a movie is, does that give me some information about what the `box_office` value could be?

**Solution:**

*This question could be answered many different ways.*

I do see some type of relationship because information about one variable does give me some idea about the other variable, so I would say there is a weak relationship/association between box office earnings and critic score.


## Question 5

Pose your own question about this data. Produce an appropriate plot to help answer your question.

**Solution:**

How does box office earnings vary by genre?

```{r}
#| fig-cap: "Box plot of movie box office earnings by genre."
#| fig-alt: "Box plots of box office earnings in millions of dollars for each movie genre for movies in 2003. The genres are western, suspense, sci-fi, rom-comedy, musical, horror, fantasy, drama, documentary, comedy, animated, adventure, action. Fantasy has the highest median box office around 100 million dollars and both musical and documentary have the lowest median box office earnings under 10 million dollars."
ggplot(movies, aes(x = box_office, y = genre)) +
  geom_boxplot(color = "darkgreen") +
  theme_bw() +
  labs(
    x = "Box office earnings (millions $)",
    y = "Genre"
  )
```

