---
title: "Data 200: Lecture 14a"
subtitle: "Presentations"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
    toc: true
    toc-depth: 1
---

```{r}
library(tidyverse)
```


# Making slides with Quarto

## Why use Quarto?

1. We are using Quarto to make the slides for our final project presentations so we can learn how to communicate our findings reproducibly.

. . .

- *Why is this important?*

. . .

2. It's simple now that we are already familiar with Quarto!

## Types of Quarto slides

We are able to make 3 different types of presentations using Quarto documents. 

- For our final projects we will be making HTML slides (called Reavel JS), just like the lecture slides!

## Let's start our presentations!

1. In your final project repo on Posit Cloud create a *Quarto Presentation* and select *Reveal JS* option (to produce html slides). Your slides should have this header/yaml:

```{}
---
title: "Untitled"
format: revealjs
---
```

2. Add an R chunk to your document after the yaml. This is where you will load all of the R packages you use
3. Render your presentation and look at your slides. Save the document as `presentation.qmd`

## 

To add a new slide you use double `##`, followed by the slide name. Add a blank line, then specify the contents of your slide

```{}
---
title: "Untitled"
format: revealjs
---

## Slide 1 title

Stuff

## Another label for next slide

More stuff
```

## 

To add header slides you can use a single `#`

```{}
---
title: "Untitled"
format: revealjs
---

# Research Question 1

## Approach

## Results

# Research Question 2

## Data filtering

##

```

. . .

You can also choose to leave a slide header blank

## yaml details

You can change the title, add a subtitle, add authors, and more!

```{}
---
title: "Investigating roller coaster speeds"
subtitle: "Team 8"
author: "Catalina Medina, Federica Ricci, and Sakshi Agarwal"
format: revealjs
---
```


## Presentation guidelines

Think about what you 

- want to show on the slides (e.g., main points, plots, hypotheses...)
- do not want to show on the slides (e.g., large blocks of text, most code, direct output, default plot axes)

In these slides you will want to use **inline R code** and **execution arguments**

# Execution options

You may have noticed some undesirable things, like messages from loading packages, or even just R code that you don't want to show.

- We can specify certain options globally (i.e., across the entire document). This is done in the yaml.
- Or we can specify options for a certain chunk. This is done at the top of the chunk


## Global arguments

When you look at our homework, we often specified that we did not want package loading messages or warnings to show in our rendered document. 

```{}
---
title: "Untitled"
format: revealjs
execute:
  echo: false
  warning: false
  message: false
---
```

- `echo` specifies if we want the code printed
- both `warning` and `message` should be set to `false` to suppress the package loading messages

## Chunk arguments

There are many different chunk arguments we can specify. I use them mostly for figures.

```{r}
#| echo: fenced
#| fig-width: 4
#| fig-height: 3
#| fig-align: center
#| fig-cap: "Penguins feature lengths by species"
#| fig-alt: "Scatterplot of bill length by flipper length, colored by species. There appears to be a positive linear relationship for each of the three species: Adelie, Chinstrap, and Gentoo."

ggplot(penguins, aes(x = flipper_len, y = bill_dep, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Flipper length (mm)",
    y = "Bill Depth (mm)",
    color = "Species"
  )
```

##

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 3
#| fig-align: center
#| fig-cap: "Penguins feature lengths by species"
#| fig-alt: "Scatterplot of bill length by flipper length, colored by species. There appears to be a positive linear relationship for each of the three species: Adelie, Chinstrap, and Gentoo."

ggplot(penguins, aes(x = flipper_len, y = bill_dep, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Flipper length (mm)",
    y = "Bill Depth (mm)",
    color = "Species"
  )
```

## Customizing

Make your slides something you are proud to share with others!

- [Quarto basics](https://quarto.org/docs/presentations/)

I encourage you to customize your slides (e.g., specifying the theme, adding images, making things appear incrementally)

- [Quarto extra capabilities](https://quarto.org/docs/presentations/revealjs/)


# Missing data

The reason behind why some data entries are missing determines what approach you should take.

- For our final project we will assume that data are missing completely at random, the reason for the missingness is not related to the variables we are measuring. For example:

  - someone forgot to answer all questions on a survey
  - data entry error occurred and was deleted
  - someone's sample was lost

## Handling missing data

Since we are assuming the data are missing completely at random, we can delete incomplete observations (aka rows with any `NA`'s).

```{r}
penguins_cleaned <- penguins |> 
  drop_na()
```

## Data cleaning

Data cleaning should be done at the top of your document without showing the code in the presentation. 

You should have a slide or two describing how you cleaned your data 

  - For example: "We dropped observations with missing data, filtered to only observations from 2007, and converted species to a factor."

## Presentation details

Please see `final-project-presentation-instructions.qmd` on CI Learn.

