---
title: "Data 200: Lecture 10a"
subtitle: "Simple Linear Regression"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
library(tidyverse)
library(broom)
library(gridExtra)

options(scipen = 999)

theme_set(theme_bw(base_size = 22))
```

## Data `babies` in `openintro` package

```{r}
library(openintro)

data("babies")
glimpse(babies)
```

##  Baby Weights


```{r}
#| output-location: column

ggplot(
  babies,
  aes(x = gestation, y = bwt)
) +
  geom_point()

```


##  Baby Weights


```{r }
#| output-location: column

ggplot(
  babies,
  aes(x = gestation, y = bwt)
) +
  geom_point() +
  geom_smooth(
    method = "lm", 
    se = FALSE
  ) 
```

`lm` stands for linear model  
`se` stands for standard error


## 

<center>


| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Birth weight | y | Response | Numeric |
| Gestation | x | Explanatory | Numeric |

</center>

## Why do we want a linear model?

. . .

Recall for the inference unit of our class, we wanted to have a model that represented the data well, so we could use that model to learn information.

. . .

A linear model is nice, when if fits the data well, because it is very understandable.


## Linear Equations Review


:::{.pull-left}

Recall from your previous math classes

$y = mx + b$

where $m$ is the slope and $b$ is the y-intercept

e.g. $y = 2x -1$
:::

. . .



:::{.pull-right}

```{r echo = FALSE, fig.height = 5, message = FALSE}
x <- c(0, 1, 2, 3, 4, 5)
y <- c(-1, 1, 3, 5, 7, 9)

as.data.frame(x = x, y = y) |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  scale_y_continuous(breaks = c(-1, 1, 3, 5, 7, 9)) +
  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5)) +
  geom_smooth(method = "lm", se = FALSE)

```

*Notice anything different between baby weights plot and this one?*
:::

##

:::{.pull-left}

**Math** class

$y = b + mx$

- $b$ is y-intercept  
- $m$ is slope  
:::


:::{.pull-left}

**Stats** class

$y_i = \beta_0 +\beta_1x_i + \epsilon_i$

- $\beta_0$ is **y-intercept**
- $\beta_1$ is **slope**  
- $\epsilon_i$ is **randomness / "random noise"**  
- $i = 1, 2, ...n$ identifier for each point
:::

. . .

Note $y_i = \beta_0 +\beta_1x_i + \epsilon_i$ is the **theoretical model**. In reality we don't have data for everything in our population, so we use a sample to get estimates for $\beta_0$ and $\beta_1$.

##

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)
```

 

We are fitting a linear regression model. 

- the `lm()` function fits a linear regression model.
- `lm` stands for linear model.
- Note that the variables are entered in y ~ x order.



## Model output

```{r}
library(broom)

tidy(linear_baby_model)
```

. . .

General formula for fitted model:

$\hat {y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$

. . .

*The hat means it is an estimate!*

. . .

**Fitted model** (Context specific):

$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464\text{ gestation}_i$

# We can use our model for prediction!

## Expected/predicted birth weight for a baby with 300 days of gestation


$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464\text{ gestation}_i$

$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464 ( 300)$

$\widehat{\text{Birth Weight}_i} =$ `r -10.1 + 0.464*300`

For a baby with 300 days of gestation the expected birth weight is `r -10.1 + 0.464*300` ounces.

## Expected/predicted birth weight for a baby with 300 days of gestation


```{r}
# Define x value for data to predict with
new_data <- data.frame(gestation = 300)

predict(linear_baby_model, newdata = new_data)
```


```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 6
#| fig-align: center

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(
    data = mutate(new_data, bwt = predict(linear_baby_model, newdata = new_data)), 
    color = "red",
    size = 2
  )
```


## Interpretation of estimates: general

- $\hat{\beta}_0$ is the predicted/expected $Y$ value when $X = 0$.

- $\hat{\beta}_1$ is the difference in predicted/expected $Y$ value associated with a 1 unit increase in $X$

## Interpretation of estimates: in context


:::{.pull-left}
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies |> 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 
```

$\hat{\beta}_1 = 0.464$ which means for one unit (day) increase in gestation period the expected increase in birth weight is 0.464 ounces.

:::

. . .

:::{.pull-right}
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies |> 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlim(0, 360) +
  ylim(-10, 180) +
  geom_abline(slope = 0.459, intercept = -8.76)
```

$\hat{\beta}_0 = -10.1$ which means for a gestation period of 0 days the expected birth weight is -10.1 ounces!!!!!!!!
(does NOT make sense)
:::



## Extrapolation

- There is no such thing as 0 days of gestation.

. . .

- Birth weight cannot possibly be -10.1 ounces.

. . .

- Extrapolation happens when we use a model outside the range of the x-values that are observed. After all, we cannot really know how the model behaves (e.g. may be non-linear) outside of the scope of what we have observed. 

# Model Error

## Baby number 148

Let's consider a baby in our data and see how far/close the true birth weight is from the predicted birth weight.

. . .

:::{.pull-left}

```{r}
babies |> 
  filter(case == 148) |> 
  select(bwt, gestation)
```

:::

:::{.pull-right}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 6

baby_148 <- babies |> 
  filter(case == 148) |> 
  select(bwt, gestation)

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red")
```


:::



## Baby #148

**Observed**

$\text{Birth Weight}_{148} =$ 160

**Expected / Predicted**

$\widehat{\text{Birth Weight}}_{148} = \hat{\beta}_0 +\hat{\beta}_1 \text{Gestation}_{148}$

$\widehat{\text{Birth Weight}}_{148} = -10.1 + 0.464\times300$

$\widehat{\text{Birth Weight}}_{148}$ = `r -10.1 + 0.464*300`



## Residual for `i = 148`

:::{.pull-left}

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red", size = 2) +
  geom_segment(
    x = 300, xend = 300, y = 128.94, yend = 160, 
    color = "red",
    size = 2,
    linetype = "dotted"
  )
```



:::

:::{.pull-right}

$\text{Birth Weight}_{148} = 160$

$\widehat{\text{Birth Weight}}_{148}$ = `r -10.1 + 0.464*300`

**Residual:**

$e_{i} = y_{i} - \hat y_{i}$

$e_{148} =$ 160 - `r -10.1 + 0.464*300` =  `r 160 -(-10.1 + 0.464*300)`


:::



## Least Squares Regression 

There are different ways we could have fit a line to our data, we choose the one ($\hat{\beta}_0$ and $\hat{\beta}_1$) that minimizes the **sum of squared residuals**:

$$e_1^2 + e_2^2 + ... + e_n^2$$

. . .

*Why didn't we choose to minimize*

$$e_1 + e_2 + ... + e_n?$$


## Inference: Confidence Interval

Recall that when we do statistics we don't just want an estimate, we want some understanding of the uncertainty of our estimate (e.g., confidence interval)

## Inference: Confidence Interval


```{r}
confint(linear_baby_model)
```

We expect a difference in birth weight between `r round(confint(linear_baby_model)["gestation", ], digits = 2)[1]` and `r round(confint(linear_baby_model)["gestation", ], digits = 2)[2]`, for a 1 day increase in gestation, with 95% confidence. 

. . .

Note that the 95% confidence interval for the slope does not contain zero and all the values in the interval are positive 

- indicating a statistically significant positive relationship between gestation and birth weight.

