---
title: "Data 200: Lecture 10a"
subtitle: "Simple Linear Regression"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
library(tidyverse)
library(broom)
library(gridExtra)

options(scipen = 999)

theme_set(theme_bw(base_size = 22))
```

## Data `babies` in `openintro` package

```{r}
library(openintro)

data("babies")
glimpse(babies)
```

##  Baby Weights


```{r}
#| output-location: column

ggplot(
  babies,
  aes(x = gestation, y = bwt)
) +
  geom_point()

```


##  Baby Weights


```{r }
#| output-location: column

ggplot(
  babies,
  aes(x = gestation, y = bwt)
) +
  geom_point() +
  geom_smooth(
    method = "lm", 
    se = FALSE
  ) 
```

`lm` stands for linear model  
`se` stands for standard error


## 

<center>


| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Birth weight | y | Response | Numeric |
| Gestation | x | Explanatory | Numeric |

</center>

## Why do we want a linear model?

. . .

Recall for the inference unit of our class, we wanted to have a model that represented the data well, so we could use that model to learn information.

. . .

A linear model is nice, when if fits the data well, because it is very understandable.


## Linear Equations Review


:::{.pull-left}

Recall from your previous math classes

$y = mx + b$

where $m$ is the slope and $b$ is the y-intercept

e.g. $y = 2x -1$
:::

. . .



:::{.pull-right}

```{r echo = FALSE, fig.height = 5, message = FALSE}
x <- c(0, 1, 2, 3, 4, 5)
y <- c(-1, 1, 3, 5, 7, 9)

as.data.frame(x = x, y = y) |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  scale_y_continuous(breaks = c(-1, 1, 3, 5, 7, 9)) +
  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5)) +
  geom_smooth(method = "lm", se = FALSE)

```

*Notice anything different between baby weights plot and this one?*
:::

##

:::{.pull-left}

**Math** class

$y = b + mx$

- $b$ is y-intercept  
- $m$ is slope  
:::


:::{.pull-left}

**Stats** class

$y_i = \beta_0 +\beta_1x_i + \epsilon_i$

- $\beta_0$ is **y-intercept**
- $\beta_1$ is **slope**  
- $\epsilon_i$ is **randomness / "random noise"**  
- $i = 1, 2, ...n$ identifier for each point
:::

. . .

Note $y_i = \beta_0 +\beta_1x_i + \epsilon_i$ is the **theoretical model**. In reality we don't have data for everything in our population, so we use a sample to get estimates for $\beta_0$ and $\beta_1$.

##

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)
```

 

We are fitting a linear regression model. 

- the `lm()` function fits a linear regression model.
- `lm` stands for linear model.
- Note that the variables are entered in y ~ x order.



## Model output

```{r}
library(broom)

tidy(linear_baby_model)
```

. . .

General formula for fitted model:

$\hat {y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$

. . .

*The hat means it is an estimate!*

. . .

**Fitted model** (Context specific):

$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464\text{ gestation}_i$

# We can use our model for prediction!

## Expected/predicted birth weight for a baby with 300 days of gestation


$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464\text{ gestation}_i$

$\widehat{\text{Birth Weight}_i} = -10.1 + 0.464 ( 300)$

$\widehat{\text{Birth Weight}_i} =$ `r -10.1 + 0.464*300`

For a baby with 300 days of gestation the expected birth weight is `r -10.1 + 0.464*300` ounces.

## Expected/predicted birth weight for a baby with 300 days of gestation


```{r}
# Define x value for data to predict with
new_data <- data.frame(gestation = 300)

predict(linear_baby_model, newdata = new_data)
```


```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 6
#| fig-align: center

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(
    data = mutate(new_data, bwt = predict(linear_baby_model, newdata = new_data)), 
    color = "red",
    size = 2
  )
```


## Interpretation of estimates: general

- $\hat{\beta}_0$ is the predicted/expected $Y$ value when $X = 0$.

- $\hat{\beta}_1$ is the difference in predicted/expected $Y$ value associated with a 1 unit increase in $X$

## Interpretation of estimates: in context


:::{.pull-left}
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies |> 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 
```

$\hat{\beta}_1 = 0.464$ which means for one unit (day) increase in gestation period the expected increase in birth weight is 0.464 ounces.

:::

. . .

:::{.pull-right}
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies |> 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlim(0, 360) +
  ylim(-10, 180) +
  geom_abline(slope = 0.459, intercept = -8.76)
```

$\hat{\beta}_0 = -10.1$ which means for a gestation period of 0 days the expected birth weight is -10.1 ounces!!!!!!!!
(does NOT make sense)
:::



## Extrapolation

- There is no such thing as 0 days of gestation.

. . .

- Birth weight cannot possibly be -10.1 ounces.

. . .

- Extrapolation happens when we use a model outside the range of the x-values that are observed. After all, we cannot really know how the model behaves (e.g. may be non-linear) outside of the scope of what we have observed. 

# Model Error

## Baby number 148

Let's consider a baby in our data and see how far/close the true birth weight is from the predicted birth weight.

. . .

:::{.pull-left}

```{r}
babies |> 
  filter(case == 148) |> 
  select(bwt, gestation)
```

:::

:::{.pull-right}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 6

baby_148 <- babies |> 
  filter(case == 148) |> 
  select(bwt, gestation)

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red")
```


:::



## Baby #148

**Observed**

$\text{Birth Weight}_{148} =$ 160

**Expected / Predicted**

$\widehat{\text{Birth Weight}}_{148} = \hat{\beta}_0 +\hat{\beta}_1 \text{Gestation}_{148}$

$\widehat{\text{Birth Weight}}_{148} = -10.1 + 0.464\times300$

$\widehat{\text{Birth Weight}}_{148}$ = `r -10.1 + 0.464*300`



## Residual for `i = 148`

:::{.pull-left}

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

babies |> 
  ggplot(aes(x = gestation, y = bwt)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red", size = 2) +
  geom_segment(
    x = 300, xend = 300, y = 128.94, yend = 160, 
    color = "red",
    size = 2,
    linetype = "dotted"
  )
```



:::

:::{.pull-right}

$\text{Birth Weight}_{148} = 160$

$\widehat{\text{Birth Weight}}_{148}$ = `r -10.1 + 0.464*300`

**Residual:**

$e_{i} = y_{i} - \hat y_{i}$

$e_{148} =$ 160 - `r -10.1 + 0.464*300` =  `r 160 -(-10.1 + 0.464*300)`


:::



## Least Squares Regression 

The goal is to minimize the **sum of squared residuals**:

$$e_1^2 + e_2^2 + ... + e_n^2$$

The estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ that we get from `lm()` are for the line the minimizes the sum of squared residuals

## Inference: Confidence Interval

Recall that when we do statistics we don't just want an estimate, we want some understanding of the uncertainty of our estimate (e.g., confidence interval)

. . .


```{r}
confint(linear_baby_model)
```

Note that the 95% confidence interval for the slope does not contain zero and all the values in the interval are positive indicating a significant positive relationship between gestation and birth weight.


# Conditions for Linear Regression

- Linearity

- Independence

- Constant Variance

- Normality of Residuals 


## Linearity conition

:::{.pull-left}

**Linear**
```{r  echo = FALSE, message = FALSE}
set.seed(84735)
x <- seq(-2, 2, by = 0.01)
y <- 4*x + 5 + rnorm(length(x), 0 , 1.5)

data_good <- data.frame(x = x, y = y) |>   
  sample_n(50)

data_good |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

:::{.pull-right}

**Non-linear**

```{r  echo = FALSE, message = FALSE}
set.seed(84735)
x <- seq(-2, 2, by = 0.01)
y <- 3*x^2 + x + 5 + rnorm(length(x), 0 , 2)

data_bad<- data.frame(x = x, y = y) |> 
  sample_n(50)

data_bad |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  stat_function(fun = function(x)3*x^2 + x + 5 )
```

:::

## Checking linearity

```{r}
#| fig-alight: center
#| fig-width: 4
#| fig-height: 3

ggplot(babies, aes(x = gestation, y = bwt)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 
```

These variables do appear to be weakly linearly related.


## Independence

Harder to check because we need to know how the data were collected.

. . .

In the description of the dataset it says _[a study]considered all pregnancies between 1960 and 1967 among women in the Kaiser Foundation Health Plan in the San Francisco East Bay area._

. . .

It is possible that babies born in the same hospital may have similar birth weight.

Correlated data examples: patients within hospitals, students within schools, people within families, time-series data. 


## Variance of residuals

:::{.pull-left}

```{r}
#| echo: false
linear_model_good <- lm(y ~ x, data = data_good)

data_good <- data_good |>
  sample_n(50) |> 
  modelr::add_residuals(linear_model_good)

linear_model_bad <- lm(y ~ x, data = data_bad)

data_bad <- data_bad |> 
  modelr::add_residuals(linear_model_bad) 
```


**Constant Variance**

```{r echo = FALSE, message = FALSE}
data_good |> 
  ggplot(aes(x = x, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

:::

:::{.pull-right}

**Non-constant variance**


```{r echo = FALSE, message = FALSE}
data_bad |> 
  ggplot(aes(x = x, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

:::

## Checking variance of residuals

```{r}
#| fig-alight: center
#| fig-width: 4
#| fig-height: 3

babies |> 
  modelr::add_residuals(linear_baby_model) |> 
  ggplot(aes(x = gestation, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue")
```

For gestation over 300 days there does appear to be less variance in residuals. This assumption does not appear to be met.


## Normality of residuals

:::{.pull-left}

**Nearly normal**

```{r echo = FALSE, message = FALSE}
data_good |> 
  ggplot(aes(x = resid)) +
  geom_density()
```

:::

:::{.pull-right}

**Not normal**


```{r echo = FALSE, message = FALSE}
data_bad |> 
  ggplot(aes(x = resid)) +
  geom_density()
```

:::

## Checking normality of residuals

```{r}
#| fig-alight: center
#| fig-width: 4
#| fig-height: 3

babies |> 
  modelr::add_residuals(linear_baby_model) |> 
  ggplot(aes(x = resid)) +
  geom_density()
```

The residuals do appear to be normally distributed.


# New example: Regressing birth weight on mother's smoking

##

Birth weight regressed on mother's smoking

<center>


| Variable | Symbol | Relationship | Type |
|---|-------------|-----------------|---------|
| Birth weight | y | Response | Numeric |
| Mother's smoking | x | Explanatory | Categorical |

</center>

. . .

```{r}
glimpse(babies)
```

## Fitting our model

```{r}
linear_baby_model2 <- lm(bwt ~ smoke, data = babies)
tidy(linear_baby_model2)
```

$\hat {y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$

$\widehat{\text{Birth Weight}_i} = \hat{\beta}_0 + \hat{\beta}_1 \text{ smoke}_i$

$\widehat{\text{Birth Weight}_i} = 123 + (-8.94\text{ smoke}_i)$

. . .

`r round(linear_baby_model2$coefficients["smoke"], digits = 2)` ounces is the expected difference in birth weight, comparing smoking mothers (`smoke` = 1) to non smoking mothers (`smoke` = 0).

## Predicting with our model

Expected bwt for a baby with a non-smoker mother

$\widehat{\text{Birth Weight}_i} = 123 + (-8.94\text{ smoke}_i)$

$\widehat{\text{Birth Weight}_i} = 123 + (-8.94\times 0)$

$\widehat{\text{Birth Weight}_i} = 123$

. . .


Expected bwt for a baby with a smoker mother

$\widehat{\text{Birth Weight}_i} = 123 + (-8.94\text{ smoke}_i)$

$\widehat{\text{Birth Weight}_i} = 123 + (-8.94\times 1)$

$\widehat{\text{Birth Weight}_i} = 114.06$

## Predicting with our model

```{r}
new_data2 <- data.frame(smoke = c(0, 1))

predict(linear_baby_model2, newdata = new_data2)
```


## Inference with our model

```{r}
confint(linear_baby_model2)
```

. . .

We estimate the expected difference in birth weight to be between `r round(confint(linear_baby_model2)["smoke",], digits = 2)[1]` and `r round(confint(linear_baby_model2)["smoke",], digits = 2)[2]` with 95% confidence, comparing smoking mothers to non smoking mothers.

. . .

Note that the confidence interval for the "slope", $\beta_1$, does not contain 0 and all the values in the interval are negative. 


## Understanding Relationships

- Just because we observe a significant relationship between $x$ and $y$, it does not mean that $x$ causes $y$. *Correlation does not prove causation!*

- Just because we observe a significant relationship in a sample that does not mean the findings will generalize to the population. 

- For these we need to understand sampling and study design. 
