---
title: "Data 200: Lecture 10b"
subtitle: "Simple linear regression modeling conditions"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

## 

```{r}
#| echo: false
options(scipen=0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)

theme_set(theme_bw(base_size = 22))
```


## Recal our `babies` data in `openintro` package

```{r}
library(openintro)

data("babies")
glimpse(babies)
```

## We fit a linear regression model

```{r}
linear_baby_model <- lm(bwt ~ gestation, data = babies)
```

## We used the model for predicting

```{r}
# Define x value for data to predict with
new_data <- data.frame(gestation = 250)

predict(linear_baby_model, newdata = new_data)
```

## We plotted the model with the data

```{r }
#| output-location: column

ggplot(
  babies,
  aes(x = gestation, y = bwt)
) +
  geom_point() +
  geom_smooth(
    method = "lm", 
    se = FALSE
  ) +
  labs(
    x = "Gestation (days)",
    y = "Birth weight (ounces)"
  )
```

## We looked at estimates from the model

```{r}
tidy(linear_baby_model)
```

- $\hat{\beta}_1$: For a 1 unit (day) increase in gestation, we expect/predict a higher birth weight of `r round(linear_baby_model$coefficients[2], digits = 2)` ounces.

- $\hat{\beta}_0$: We expect/predict babies with a gestation of 0 days to have a birth weight of `r round(linear_baby_model$coefficients[1], digits = 2)`....

. . .

$\beta_0$ *typically doesn't have meaning in context so we typically don't worry about interpreting it.*


## We used the model for confidence intervals 

```{r}
confint(linear_baby_model, level = 0.95)
```

95% Confidence Interval for $\beta_1$

For a 1 unit (day) increase in gestation, we expect a higher birth weight between 0.41 and 0.52 ounces, with 95% confidence.


# Why we do linear regression

Let $Y$ the response variable (outcome of interest) and $X$ be the explanatory variable (explains changes in $Y$)

- We could use more complicated models for $Y$ and $X$, but more complicated typically means less understandable

- Linear regression is easily understandable

## Simple linear regression purpose 1: 


**Prediction** (aka "machine learning")

- Predict expected value of $Y$, using $X$:
  
  - To use our model on future data where only $X$ is known

## Simple linear regression purpose 2: 

**Inference**

- Confidence interval for $\beta_1$: 

  - To learn about the relationship between $Y$ and $X$

. . .  

- Hypothesis test for $\beta_1$: (*Not covered*)

  - To test if there is a linear relationship between $Y$ and $X$ (numeric $X$)
  - To test if the mean of $Y$ differs for different $X$ (categorical $X$)



# Conditions for Simple Linear Regression

Whether we do prediction or inference, we need to make sure our model is appropriate:

- Independence

- Linearity

- Constant Variance

- Normality of Residuals 

## Checking conditions for simple linear regression

- Independence (need context)

- Linearity (plot residual)

- Constant Variance (plot residuals)

- Normality of Residuals (plot residuals)

## Independence

Independence of observations means knowing one observation should not give me information about another observation.

Correlated data examples: 

- Time-series data (aka collecting multiple observations on the same person)

  - If I wanted to learn about typical body temperature, I wouldn't want to collect multiple measurements from one person

## Independence

Correlated data examples continued: 

- People within the same families 

  - If I wanted to estimate the proportion of families that eat dinner together, I wouldn't record observations from people within the same family

- Students within the same class
- Patients from the same hospital


## Checking Independence

Harder to check because we need to know how the data were collected.

. . .

In the description of the dataset it says: 

> \[a study\] considered all pregnancies between 1960 and 1967 among women in the Kaiser Foundation Health Plan in the San Francisco East Bay area.

. . .

If they all had the same doctor I would think this condition was not met

- Sometimes this requires a little research into the area to determine

## Linearity conition

Recall residuals are the errors: $e_i = y_i - \hat{y}_i$ 

Vocabulary: $\hat{y}$ = predicted $y$ = "fitted $y$"

. . .

```{r}
#| echo: false
# Simulate data

set.seed(84735)

sample_size <- 100

x <- seq(-2, 2, by = 0.01)
y <- 4 * x + 5 + rnorm(length(x), mean = 0 , sd = 1.5)

data_good <- data.frame(x = x, y = y) |>   
  sample_n(sample_size)

linear_model_good <- lm(y ~ x, data = data_good)

data_good <- data_good |>
  add_residuals(linear_model_good)

x <- seq(-2, 2, by = 0.01)
y <- 3 * x^2  + 5 + rnorm(length(x), mean = 0 , sd = 2)

data_bad <- data.frame(x = x, y = y) |> 
  sample_n(sample_size)

linear_model_bad <- lm(y ~ x, data = data_bad)

data_bad <- data_bad |>
  add_residuals(linear_model_bad)
```


:::{.pull-left}

**Linear**
```{r}
#| echo: false
# data_good |> 
#   add_residuals(linear_model_good) |> 
#   ggplot(aes(x = y, y = resid)) +
#   geom_point() +
#   geom_hline(yintercept = 0, color = "purple", linetype = "dashed")

fortify(linear_model_good) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

Random => looks good

:::

:::{.pull-right}

**Non-linear**

```{r}
#| echo: false
# data_bad |> 
#   add_residuals(linear_model_bad) |> 
#   ggplot(aes(x = y, y = resid)) +
#   geom_point() +
#   geom_hline(yintercept = 0, color = "purple", linetype = "dashed")

fortify(linear_model_bad) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

Pattern => Uncaptured pattern

:::

## Checking linearity

```{r}
fortify(linear_baby_model) |> # Extracts residuals and predicted/fitted y
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

Looks generally random, so linearity condition appears to be met.



## Variance of residuals

```{r}
#| echo: false
x <- seq(-2, 2, by = 0.01)
y <- 4 * x + 5 + rnorm(length(x), 0 , 1.5)

data_good <- data.frame(x = x, y = y) |>   
  sample_n(sample_size)

linear_model_good <- lm(y ~ x, data = data_good)

data_good <- data_good |>
  add_residuals(linear_model_good)

x <- seq(-2, 2, by = 0.01)
y <- 3 * x^2 + x + 5 + rnorm(length(x), mean = 0, sd = 2 * x^2)

data_bad <- data.frame(x = x, y = y) |> 
  sample_n(sample_size)

linear_model_bad <- lm(y ~ x, data = data_bad)
```


:::{.pull-left}


**Constant Variance**

```{r echo = FALSE, message = FALSE}
fortify(linear_model_good) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

:::

:::{.pull-right}

**Non-constant variance**


```{r echo = FALSE, message = FALSE}
fortify(linear_model_bad) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

:::

## Checking variance of residuals

```{r}
#| fig-alight: center

fortify(linear_baby_model) |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point()
```

The variance of the residuals does appear to be relatively constant. Nothing too concerning.


## Normality of residuals

:::{.pull-left}

**Nearly normal**

```{r echo = FALSE, message = FALSE}
fortify(linear_model_good) |> 
  ggplot(aes(x = .resid)) + 
  geom_density()
```

:::

:::{.pull-right}

**Not normal**


```{r echo = FALSE, message = FALSE}
fortify(linear_model_bad) |> 
  ggplot(aes(x = .resid)) + 
  geom_density()
```

:::

## Checking normality of residuals

```{r}
#| fig-alight: center
#| fig-width: 4
#| fig-height: 3

fortify(linear_baby_model) |> 
  ggplot(aes(x = .resid)) + 
  geom_density()
```

The residuals do appear to be normally distributed.

# Week 10 lab question 9


