---
title: "Data 200"
subtitle: "Summarizing Major Points From Lecture"
format: 
  html:
    toc: true
    toc-expand: true
    toc-depth: 3
knitr:
  opts_chunk:
    message: false
    warning: false
---

We are reaching the end of the coding focused unit of this course, so I think it is important to summarize what we have covered. 
The lectures and labs provided a lot of examples, but sometimes a lot of information can make it harder to see the big picture.

*I recommend you provide an example for most of these.*
*I do not just expect you to know the functions we discussed in class, I also expect you to know how to use them.*

## Lecture 1b: Toolkit

What coding language are we using?

What software/app are we using to code?

What type of document are we writing our code in?

What is the difference between writing code in the Console versus writing code in a Quarto file?


### Coding

How can I find help documentation?

Say I am trying to use a function called `pretend_function()`. 
What is likely the issue if I get an error message saying:  
"Error in pretend_function() : could not find function 'pretend_function'".

How do I load a package?

How do I use inline code?

How can I view what is in an object?

What are the two main rules for naming objects according to the tidyverse style guide?


## Lecture 2b: Describing data with numbers

### Data set structure

What does 1 row in a dataset represent?

How can I compute the number of rows?

What does 1 column in a dataset represent?

How can I compute the number of columns?

What is a function I can use to view the structure of a data frame and some of its contents?


### Variables

What are the two main types of variables and what are each of their subtypes?

What are appropriate R types for each variable type?

What are summary statistics I can compute for categorical variables?

What are summary statistics I can compute for numeric variables? Which ones tell me about a typical value and which inform me about how much my data vary?


## Lecture 3a: Plot types

Which plots are good for visualizing:

- One categorical variable
- One numeric variable
- Two categorical variables
- Two numeric variables
- One numeric and one categorical
- Two numeric and one categorical

Specify the plotting function for each plot above (i.e. for a bar plot we use `geom_bar()`).

## Lecture 3b: Improving plots

What is the difference between specifying `color = ` in the `aes()` function versus in the plot function? Consider the following example:
```{r}
library(tidyverse)
data(penguins)

ggplot(penguins, aes(x = bill_len, y = body_mass, color = species)) +
  geom_point()
```

```{r}
ggplot(penguins, aes(x = bill_len, y = body_mass)) +
  geom_point(color = "red")
```

Fill in the blank: To follow tidyverse style when plotting, each _____ should be on its own line. 
If we want to add another layer to a plot, we add _____ to the end of the previous line of code. 
The options are: "layer", "argument", "aesthetic", "|>", "+", "()".

How do I specify the labels for plot aesthetics such as axes, color, title, caption, alternate text.

What is the difference between a plot title and a plot caption?

What is one thing we can do to make visualizations more accessible to people with a visual impairment?


## Lecture 4a: Data wrangling part 1

### Piper operator

What is the symbol for the pipe operator and why is it helpful?

If I have several rows of data wrangling piping, how do I save the new data frame?


### Subsetting columns

What function can I use to subset columns from a data frame?

How would I specify I want to only keep two columns from a data frame?

How would I specify I want to drop a column from a data frame? 


### Subsetting rows

If I know the row numbers for observations I want to drop, what function should I used

If I was to subset rows that meet a certain condition, what function should I use?

What are the 6 relational operators we discussed in class and what do each of them mean?

What are the two logical operators we discussed in class and what do each of them mean?


### Changing variables

What function can I use to define a new variable/column for my data frame?

What function can I use to alter an existing variable/column from my data frame?

What function can I use when defining/altering a variable to specify different values based on conditions? 
Think about the `base_pay_level` variable we defined in lecture.

If I wanted to convert a variable in a data frame from an integer R type to a character R type, how would I do that?

What function will convert the variable names in my data frame to comply with the tidyverse style guide naming convention?


## Lecture 4b: Data wrangling part 2

### Aggregating data

If I want to use the `summarize()` function to calculate some summary statistics for different levels of a categorical variable, what function will I need to use before the `summarize()` function?


### Joining data

The join functions help us merge two data frames using columns both data sets have in common.
What function should I use if I want my newly joined data frame to contain all rows and columns present in either of the original data frames?

What function should I use if I want my newly joined data frame to only contain all rows and columns present in both of the original data frames?

What function should I use if I want my newly joined data frame to only contain all rows and columns present in the first original data frame that were not present in the second original data frame?

What function should I use if I want my newly joined data frame to only contain all rows and columns present in the first original data frame, and all of the columns present in the second original data frame?


## Lecture 5a: Data wrangling part 3

### Factors

What function can we use instead of `as.factors()` to convert a variable to a factor?

How can we specify the order of levels for a factor?

How can we rename the levels of a factor?


### Strings

What package provides functions for manipulating strings (aka factors and character R types)?


### Dates

What package provides functions for manipulating dates?

What are R types that are appropriate for variables that contain date and or time information?

