---
title: "Data 200: Lab 2b Solutions"
subtitle: "Describing data with numbers"
author: "Professor Catalina Medina"
format: html
---

*Please always start labs and homework by changing the author to your name. Please answer questions using complete sentences where appropriate.*

# Part 1: Introduction to the data

In this lab we will be working with Wii Mario Kart Ebay auction data.

```{r}
library(tidyverse)
library(openintro)

data("mariokart")
```

The `mariokart` data is in the `openintro` package.

### Question 1. 

Please look at the data documentation using `?mariokart` in the console.
For each variable below please specify what type of variable it is and an appropriate R variable type. The first one is done for you as an example.

**Solution:**

- `duration`: Numeric discrete (numeric or integer)
- `id`: Categorical nominal (character or factor)
- `n_bids`: Numeric discrete (numeric or integer)
- `cond`: Categorical nominal (character or factor)
- `total_pr`: Numeric continuous (numeric or double)
- `title`: Categorical nominal (character or factor)


### Question 2. 

Please view the data and make sure the R variable types for each column are appropriate for the variable types. For example the variable `start_pr` is numeric and the R variable column type in the data is "dbl" meaning `double`. A `double` is just a numeric that can take decimals, so it is appropriate.

Note any variables that have an inappropriate R type.

**Solution:**

```{r}
glimpse(mariokart)
```

The `id` variable is actually categorical, but the column type is "double". 


# Part 2: Summarizing our data numerically

I'm curious how much people ended up selling the game for.

### Question 3. 

Which variable contains this information?


**Solution:**

The variable `total_pr`.

### Question 4. 

Please calculate a value that tells me how much the game typically sold for. Use inline code to answer. 

**Solution:**

Two options we discussed for demonstrating "typical" game sales are mean and median, since they are both measures of centrality.

*When you render inline code you may see a lot of decimals printed. When talking about price it makes sense to only display two decimal places. If you save you calculation into an object called `typical_price` then you can use `round(typical_price, digits = 2)`*

```{r}
total_price_mean <- summarize(mariokart, mean(total_pr))
total_price_mean <- round(total_price_mean, digits = 2)
```

The mean total price was $`r total_price_mean`.


### Question 5. 

Please calculate a value that describes the variability in final sale price.

**Solution:**

Two options we discussed for demonstrating "variability" in game sales are variance and standard deviation. 
Standard deviation is easier to understand because it is in the same units as the original data.

```{r}
total_price_sd <- summarize(mariokart, sd(total_pr))
total_price_sd <- round(total_price_sd, digits = 2)
```

The standard deviation of total price was $`r total_price_sd`.


## Question 6

```{r}
janitor::tabyl(mariokart, var1 = cond)
```

What percent of games were sold new?

**Solution:**

41.3% of games were sold new.

