---
title: "Data 200: Lab 2b"
subtitle: "Familiarizing with data"
author: "Professor Catalina Medina"
format: html
---

*Please always start labs and homework by changing the author to your name. Please answer questions using complete sentences where appropriate.*

# Part 1: Introduction to the data

In this lab we will be working with Wii Mario Kart Ebay auction data.

```{r}
library(tidyverse)
library(openintro)

data("mariokart")
```

The `mariokart` data is in the `openintro` package.

## Question 1 

Please look at the data documentation using `?mariokart` in the console.
For each variable below please specify what type of variable it is and an appropriate R variable type. The first one is done for you as an example.

**Solution:**

- `duraton`: Numeric discrete
- `id`: 
- `n_bids`: 
- `cond`:
- `total_pr`: 
- `title`: 


## Question 2 

Please view the data and make sure the R variable types for each column are appropriate for the variable types. For example the variable `start_pr` is numeric and the R variable column type in the data is "dbl" meaning `double`. A `double` is just a numeric that can take decimals, so it is appropriate.

Note any variables that have an inappropriate R type.


# Part 2: Summarizing our data numerically

I'm curious how much people ended up selling the game for.


## Question 3 

Which variable contains this information?


## Question 4 

Please calculate a value that tells me how much the game typically sold for. Use inline code to answer. 


## Question 5 

Please calculate a value that describes the variability in final sale price. Use inline code to answer. 


## Question 6

```{r}
janitor::tabyl(mariokart, cond)
```

What percent of games were sold new?

